---
title: "Enzymes - GRaMPS"
author: "Heather Throop"
format: html
editor: visual
---

## Overview

This is a compilation, cleaning, analysis, and plotting of enzyme activity data from Brittney Monus's MS research in the GRaMPS project.

Code was written by Heather Throop, streamlining and updating the original analyses that Brittney did for her MS.

Written:\
2025-10-08 Heather Throop

## SetUp

```{r}
#| label: load packages
library(tidyverse)
library(bestNormalize)
library(cowplot)
library(grid)
library(here)
library(car)
library(purrr)
library(tibble)
```

## Ingest and Prepare Data

```{r}
#| label: read data

#Enzyme analyses
Enzymeslong = read.csv(here("data","EnzymeResultslong.csv"))

#Change grouping variables to factors
Enzymeslong <- Enzymeslong |>
  mutate(Enzyme = factor(Enzyme),
         Site = factor(Site),
         Treatment = factor(Treatment),
         Microsite = factor(Microsite))

# Order the variables
Enzymeslong$SiteOrder = 
  factor(Enzymeslong$Site, 
         levels=c("BP","ANT","BC","ARB","CC"), 
         labels=c("Desert Scrub","Desert Grassland","Juniper Savanna","Ponderosa Pine Meadow","Mixed Conifer Meadow")) 
Enzymeslong$EnzymeOrder = 
  factor(Enzymeslong$Enzyme, 
         levels=c("BG","CBH","NAG","PHOS"), 
         labels=c("Beta Glucosidase","Cellobiohydrolase","N-Acetyl glucosaminidase","Phosphatase")) 
Enzymeslong$TreatmentOrder = 
  factor(Enzymeslong$Treatment, 
         levels=c("Exl","Ctrl","Add"), 
         labels=c("Exclusion","Ambient","Addition")) 

```

```{r}
#| label: explore data

#summary 
EnzymeMeans <- Enzymeslong |>
  group_by(Site, Treatment, Enzyme) |>
  summarise(
    N = sum(!is.na(Activity)),
    mean_activity = mean(Activity, na.rm = TRUE),
    sd_activity = sd(Activity, na.rm = TRUE),
    se_activity = sd_activity / sqrt(N),
    .groups = "drop"
  )
print(EnzymeMeans)

# visualize the distribution
ggplot(Enzymeslong, aes(x = Activity, fill = Treatment)) +
  geom_density(alpha = 0.4) +
  facet_wrap(~ Enzyme, scales = "free") +
  theme_bw()
```

```{r}
#| label: assess outliers

# Visualize potential outliers by treatment or site
ggplot(Enzymeslong, aes(x = Treatment, y = Activity, fill = Treatment)) +
  geom_boxplot(outlier.color = "red") +
  facet_wrap(~ Enzyme, scales = "free") +
  theme_bw() +
  theme(legend.position = "none")
# There are a few outliers (2 BG, 6 CBH, 1 NAG, 2 PHOS), all of which are high. 

# Remove outliers based on 1.5*IQR rule within each group
Enzymes_outliers_rmvd <- Enzymeslong |>
  group_by(Enzyme, Site, Treatment) |>
  filter(between(
    Activity,
    quantile(Activity, 0.25, na.rm = TRUE) - 1.5 * IQR(Activity, na.rm = TRUE),
    quantile(Activity, 0.75, na.rm = TRUE) + 1.5 * IQR(Activity, na.rm = TRUE)
  )) |>
  ungroup()
```

## ANOVAs

```{r}
#| label: check ANOVA assumptions

# Transform the activity variable if needed and create summary table
summary_list <- list()  # store summary per enzyme

# Loop over each enzyme for transformation + plots
for(enzyme_name in levels(Enzymes_outliers_rmvd$Enzyme)){
  
  # Subset enzyme
  df_enzyme <- Enzymes_outliers_rmvd |> filter(Enzyme == enzyme_name)
  
  # Skip if too few points
  if(nrow(df_enzyme) < 3){
    warning(paste("Skipping", enzyme_name, "- not enough data"))
    next
  }
  
  # --- Normality check and transform if needed ---
  shapiro_raw <- shapiro.test(df_enzyme$Activity)
  
  if(shapiro_raw$p.value < 0.05){
    bn_obj <- bestNormalize(df_enzyme$Activity, allow_orderNorm = FALSE)
    df_enzyme <- df_enzyme |> mutate(Activity_trans = predict(bn_obj, newdata = Activity))
    transform_used <- class(bn_obj$chosen_transform)[1]
    shapiro_trans <- shapiro.test(df_enzyme$Activity_trans)
  } else {
    df_enzyme <- df_enzyme |> mutate(Activity_trans = Activity)
    transform_used <- "None"
    shapiro_trans <- shapiro_raw
  }
  
  # --- Levene's Test ---
  df_enzyme <- df_enzyme |> mutate(Group = interaction(Site, Treatment, drop = TRUE))
  levene_result <- leveneTest(Activity_trans ~ Group, data = df_enzyme)
  
  # --- Diagnostic plots ---
  par(mfrow = c(2,2), oma = c(0,0,3,0))  # 2x2 layout with outer title
  hist(df_enzyme$Activity, main = "Raw Activity", xlab = "Activity", col = "grey")
  qqnorm(df_enzyme$Activity, main = "Q-Q Plot (Raw)"); qqline(df_enzyme$Activity)
  hist(df_enzyme$Activity_trans, main = "Transformed Activity", xlab = "Activity_trans", col = "lightblue")
  qqnorm(df_enzyme$Activity_trans, main = "Q-Q Plot (Transformed)"); qqline(df_enzyme$Activity_trans)
  mtext(paste("Enzyme:", enzyme_name), outer = TRUE, line = 1.5, font = 2, cex = 1.2)
  par(mfrow = c(1,1))  # reset layout
  
  # --- Update the main dataframe with Activity_trans ---
  Enzymes_outliers_rmvd$Activity_trans[Enzymes_outliers_rmvd$Enzyme == enzyme_name] <- df_enzyme$Activity_trans
  
  # --- Store summary ---
  summary_list[[enzyme_name]] <- tibble(
    Enzyme = enzyme_name,
    variable = "Activity_trans",
    transformed = transform_used,
    shapiro_p_raw = shapiro_raw$p.value,
    shapiro_p_trans = shapiro_trans$p.value,
    levene_p = levene_result$`Pr(>F)`[1]
  )
}

# Combine summary table
summary_table <- bind_rows(summary_list)
summary_table


```

```{r}
#| label: 2-way ANOVA

# Define ANOVA terms
anova_terms <- "Site*Treatment"  # main effects + interaction

# Output file
output_file <- here("results", "anova_summary_Enzymes.txt")
sink(output_file)

cat("=== Transformation Summary ===\n")
print(summary_table)

# Loop over each Enzyme
for(enzyme_name in levels(Enzymes_outliers_rmvd$Enzyme)){
  
  cat("\n=============================\n")
  cat("Enzyme:", enzyme_name, "\n")
  cat("=============================\n\n")
  
  # Subset enzyme
  df_enzyme <- Enzymes_outliers_rmvd |> filter(Enzyme == enzyme_name)
  
  # Two-way ANOVA
  anova_model <- aov(as.formula(paste("Activity_trans ~", anova_terms)), data = df_enzyme)
  
  # ANOVA summary
  anova_summary <- summary(anova_model)
  print(anova_summary)
  
  # Pairwise comparisons (Tukey HSD)
  tukey_results <- TukeyHSD(anova_model)
  cat("\n--- Pairwise Comparisons (Tukey's HSD) ---\n")
  for(term in names(tukey_results)){
    cat("\n---", term, "---\n")
    print(tukey_results[[term]])
  }
  
  cat("\n\n")
}

sink()  # close the output file
```

## Boxplots

```{r}
#| label: boxplot

# Create a vector of letters for facet labels
enzyme_levels <- unique(Enzymes_outliers_rmvd$EnzymeOrder)
facet_labels <- paste0(LETTERS[1:length(enzyme_levels)], ". ", enzyme_levels)

# Create a lookup table for annotations
label_df <- tibble(
  EnzymeOrder = enzyme_levels,
  label = facet_labels
)

# Create boxplot with facet labels in upper-left
plot_Enzyme <- ggplot(data = Enzymes_outliers_rmvd, 
                      aes(x = SiteOrder, y = Activity, fill = TreatmentOrder)) +
  geom_boxplot(position = position_dodge(0.8)) +
  facet_grid(
    vars(EnzymeOrder),
    scales = "free"
  ) +
  scale_fill_manual(
    name = NULL,
    values = c("#de2d26", "#bdbdbd", "#08519c")
  ) +
  labs(
    x = "Site",
    y = expression(paste(Enzyme~Activity~"(nmol " * h^{-1} * " " * g^{-1} * " soil)"))
  ) +
  scale_x_discrete(labels = label_wrap_gen(width = 10)) +
  theme_grey(base_size = 17) +
  theme(
    strip.background = element_blank(),  # removes grey facet boxes
    strip.text = element_blank() ) +        # removes facet labels
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.ticks.x = element_blank(),
    legend.position = c(0.1, 0.65),
    panel.background = element_blank(),
    panel.border = element_rect(color = "black", fill = NA)
  ) +
  # Add facet labels in upper-left of each panel
  geom_text(
    data = label_df,
    aes(x = -Inf, y = Inf, label = label),
    inherit.aes = FALSE,
    hjust = -0.1, vjust = 1.2,
    size = 5, fontface = "plain" ) +
  theme(legend.position = c(0.1, 0.1))
plot_Enzyme

ggsave(
  filename = here("figures", "Enzyme_boxplot.png"),  
  plot = plot_Enzyme,
  width = 12,       # width in inches
  height = 8,       # height in inches
  dpi = 300         # high-resolution PNG
)

```

There are significant site effects for all enzymes (and no other significant effects), so summarize the activity values for each site to use those values in comparisons.

```{r}
#| label: compare site means

#summary 
SiteMeans <- Enzymes_outliers_rmvd |>
  group_by(Enzyme, SiteOrder) |>
  summarise(
    N = sum(!is.na(Activity)),
    mean_activity = mean(Activity, na.rm = TRUE),
    sd_activity = sd(Activity, na.rm = TRUE),
    se_activity = sd_activity / sqrt(N),
    .groups = "drop"
  )
print(SiteMeans)
```
