---
title: "Monus_GRaMPS"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

# Overview

This code works up and analyzes most data from Brittney Monus's MS thesis on the GRaMPS plots. This code was compiled by Heather Throop, based largely on the code written originally by Brittney Monus.

Data, graphing, and analysis of ANPP and root biomass data.

## Initial Set Up

```{r}
library(tidyverse)
library(here)
library(bestNormalize)
library(cowplot)
```

# ANPP and Root Biomass

## ANPP

### Calculations

```{r}
# load in 2018 ANPP data
ANPP_df <- read_csv(here("data","ANPP.csv"), show_col_types = FALSE)

# Change treatment names to align with Monus data sets
ANPP_df <- ANPP_df |>
    mutate(
    Treatment = as.factor(Treatment) |> 
      recode_factor(C = "Ctrl", D = "Exl", W = "Add"))

# calculate canopy cover for each Observation 
# From Seth "To calculate total vegetation cover per quadrat, multiply the canopy cover value by the count, then sum all these products for every species in the quadrat....It sounds like Brittney used plot as a level of replication, so could take average of the 2 quadrats in each plot. "
ANPP_df <- ANPP_df |>
    mutate(
    Canopy_coverXcount = Canopy_cover * Count)

# sum Canopy_coverXcount and ANPP for each species within individual quadrats
# note that per HT's conversation (2024-03-27) with SM, ANPP already has 'count' taken into account, no need to multiple it
ANPP_by_spp_by_quadrat <- ANPP_df |>
  group_by(Site, Treatment, Plot, Quadrat, Species) |>
  summarize(
    CanopyCoverBySpp = sum(Canopy_coverXcount, na.rm=TRUE),
    ANPPBySpp = sum(NPP_g_m2, na.rm=TRUE)) 

# sum canopy cover and ANPP data for all species within each quadrat
ANPP_by_quadrat <- ANPP_by_spp_by_quadrat |>
  group_by(Site, Treatment, Plot, Quadrat) |>
  summarize(
    QuadCanopyCover = sum(CanopyCoverBySpp, na.rm=TRUE),
    QuadANPP = sum(ANPPBySpp, na.rm=TRUE)) 

# reduce cover and ANPP data to means for each plot
ANPP_by_plot <- ANPP_by_quadrat |>
  group_by(Site, Treatment, Plot) |>
  summarize(
    plot_mean_Cover = mean(QuadCanopyCover, na.rm=TRUE),
    plot_mean_ANPP = mean(QuadANPP, na.rm=TRUE),
    n_quadrats    = n())

# reduce ANPP and cover data to means for site x treatment combination
ANPP_by_trt <- ANPP_by_plot |>
  group_by(Site, Treatment) |>
  summarize(
    n_plots = n(),
    mean_ANPP = mean(plot_mean_ANPP, na.rm=TRUE),
    sd_ANPP = sd(plot_mean_ANPP),
    se_ANPP = sd_ANPP/sqrt(n_plots),
    mean_Cover = mean(plot_mean_Cover, na.rm=TRUE),
    sd_Cover = sd(plot_mean_Cover),
    se_Cover = sd_Cover/sqrt(n_plots)
    )

# quick plot to look at mass versus cover
plot(x = ANPP_by_trt$mean_ANPP, y = ANPP_by_trt$mean_Cover)
```

## Root Biomass (g m-2)

### Calculations

```{r}
# load in root biomass data
root_biomass_df <- read_csv(here("data","RootBiomass.csv"), show_col_types = FALSE)

# Convert from long to wide format for root core depth
root_biomass_wide_df <- root_biomass_df |>
  pivot_wider(names_from = Depth, values_from = Root_mass,
              names_prefix = "Root_mass_")

# do a second pivot wider to separate out the two microsites
root_biomass_wide_df <- root_biomass_wide_df |>
  pivot_wider(names_from = Microsite, values_from = c(Root_mass_upper,
                                                      Root_mass_lower),
              names_sep = "_")

# Add core depths together 
# Calculate roots biomass in units of g m-2, based on the 7 cm diameter corer
root_biomass_wide_df <- root_biomass_wide_df |>
    mutate(
    Root_mass_Can = Root_mass_upper_Can + Root_mass_lower_Can,
    Root_mass_IC = Root_mass_upper_IC + Root_mass_lower_IC,
    Root_mass_Can_gm2 = Root_mass_Can /((3.14159*3.5^2)/(100*100)),
    Root_mass_IC_gm2 = Root_mass_IC / ((3.14159*3.5^2)/(100*100))
    )

```

### 

## Combined ANPP and Root Biomass

```{r}
# group by depth and canopy
ANPPandRoot_df <- left_join(ANPP_by_plot, root_biomass_wide_df, by = c("Site", "Treatment", "Plot"))

# calculate area-weighted root biomass
ANPPandRoot_df <- ANPPandRoot_df |>
    mutate(
    Root_mass_Can_gm2_weighted = Root_mass_Can_gm2 * plot_mean_Cover/100,
    Root_mass_IC_gm2_weighted = Root_mass_IC_gm2 * (1-plot_mean_Cover/100),
    Root_mass_total_gm2 = Root_mass_Can_gm2_weighted + Root_mass_IC_gm2_weighted,
    root_to_ANPP = Root_mass_total_gm2/plot_mean_ANPP
    )

# quick plot to look at mass versus cover
plot(x = ANPPandRoot_df$Root_mass_total_gm2, y = ANPPandRoot_df$plot_mean_ANPP)

# reduce ANPP and root biomass to means for site x treatment combination
ANPP_roots_by_trt <- ANPPandRoot_df |>
  group_by(Site, Treatment) |>
  summarize(
    n_plots = sum(!is.na(plot_mean_ANPP)),
    mean_ANPP = mean(plot_mean_ANPP, na.rm=TRUE),
    sd_ANPP = sd(plot_mean_ANPP, na.rm=TRUE),
    se_ANPP = sd_ANPP/sqrt(n_plots),
    n_roots = sum(!is.na(Root_mass_total_gm2)),
    mean_Root_gm2 = mean(Root_mass_total_gm2, na.rm=TRUE),
    sd_Root_gm2 = sd(Root_mass_total_gm2, na.rm=TRUE),
    se_Root_gm2 = sd_Root_gm2/sqrt(n_roots)
    )

```

### Plots - ANPP and root biomass by treatment and site

#### Set up legends

```{r}
#|label: export L2 data

# ANPP Plots 
ANPPandRoot_df <- ANPPandRoot_df |>
  mutate(
    SiteOrder = factor(Site, levels=c("BP","ANT","BC","ARB","CC"), 
                     labels=c("Desert Scrub","Desert Grassland","Juniper Savanna",
                              "Ponderosa Pine Meadow","Mixed Conifer Meadow")),
    TreatmentOrder = factor(Treatment, levels=c("Exl","Ctrl","Add"), 
                        labels=c("Exclusion","Ambient","Addition")) 
  )


# export L2 data as csv
write.csv(ANPPandRoot_df, here("data", "calculated_outputs", "ANPPandRootBiomass_L2.csv"))
```

#### ANPP

Note: Analyses can be started here with the L2 data saved in last chunk

```{r}
#| label: ANPP-plot

# Define site labels and levels
site_labels <- c(
  "Desert Scrub" = "Desert Scrub",
  "Desert Grassland" = "Desert Grassland",
  "Juniper Savanna" = "Juniper Savanna",
  "Ponderosa Pine Meadow" = "Ponderosa Pine\nMeadow",
  "Mixed Conifer Meadow" = "Mixed Conifer\nMeadow"
)
site_levels <- names(site_labels)

ANPPandRoot_df <- ANPPandRoot_df |>
  mutate(SiteOrder = factor(SiteOrder, levels = names(site_labels)))

ANPP <- ANPPandRoot_df |>
  ggplot(aes(x = SiteOrder, y = plot_mean_ANPP, fill = TreatmentOrder)) + 
  geom_boxplot(position = position_dodge(0.8)) +  
  scale_fill_manual(name = NULL, values = c("#de2d26", "#bdbdbd", "#08519c")) +
  scale_y_continuous(limits = c(0, 220), expand = c(0, 0)) +
  scale_x_discrete(labels = site_labels) +
  labs(
   x = "Site", 
   y = expression("Mean Plot ANPP (g " * m^{-2} * ")")) +
  theme_grey(base_size = 18) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.ticks.x = element_blank(),
  #  axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = c(0.3, 0.7),
    panel.background = element_blank(),
    panel.border = element_rect(color = "black", fill = NA)
  ) 
ANPP

ggsave(here("figures", "ANPP.png"), plot = ANPP, width = 10, height = 8)

```

#### Root biomass - Canopy

```{r}
#| label: boxplot-root-biomass-canopy
roots_Can_gm2 <- ANPPandRoot_df |>
  ggplot(aes(x = SiteOrder, y = Root_mass_Can_gm2, fill = TreatmentOrder)) + 
  geom_boxplot(position = position_dodge(0.8)) +  
  scale_fill_manual(name = "", values = c("#de2d26", "#bdbdbd", "#08519c")) +
#  scale_y_continuous(limits = c(0, 6), breaks = c(0, 2, 4, 6)) +
  labs(x = "Site", 
       y = "Root Mass in Canopy microsites (g m-2)") +  
  theme_grey(base_size = 18) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = c(0.3, 0.7),
    panel.background = element_blank(),
    panel.border = element_rect(color = "black", fill = NA)
  ) 
roots_Can_gm2

ggsave(here("figures", "roots_Can_gm2.png"), plot = roots_Can_gm2, width = 10, height = 8)

```

#### Root biomass - Intercanopy

```{r}
#|label: column-graph-rootbiomass-IC-by-trt-and-site

roots_IC_gm2 <- ANPPandRoot_df |>
  ggplot(aes(x = SiteOrder, y = Root_mass_IC_gm2, fill = TreatmentOrder)) + 
  geom_boxplot(position = position_dodge(0.8)) +  
  scale_fill_manual(name = "Treatment", values = c("#de2d26", "#bdbdbd", "#08519c")) +
#  scale_y_continuous(limits = c(0, 6), breaks = c(0, 2, 4, 6)) +
  labs(x = "Site", 
       y = "Root Mass in IC microsites (g m-2)") +  
  theme_grey(base_size = 18) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = c(0.3, 0.7),
    panel.background = element_blank(),
    panel.border = element_rect(color = "black", fill = NA)
  ) 
roots_IC_gm2

ggsave(here("figures", "roots_IC_gm2.png"), plot = roots_IC_gm2, width = 10, height = 8)
```

#### Root biomass - Combined

```{r}
#|label: column-graph-rootbiomass-total-by-trt-and-site

roots_total_gm2 <- ANPPandRoot_df |>
  ggplot(aes(x = SiteOrder, y = Root_mass_total_gm2, fill = TreatmentOrder)) + 
  geom_boxplot(position = position_dodge(0.8)) +  
  scale_fill_manual(name = "", values = c("#de2d26", "#bdbdbd", "#08519c")) +
  scale_y_continuous(limits = c(0, 1100), expand = c(0, 0)) +
 labs(
   x = "Site", 
   y = expression("Root Mass (g " * m^{-2} * ")")) +
  scale_x_discrete(labels = site_labels) +
  theme_grey(base_size = 18) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.ticks.x = element_blank(),
    legend.position = c(0.3, 0.7),
    panel.background = element_blank(),
    panel.border = element_rect(color = "black", fill = NA)
  ) 
roots_total_gm2

ggsave(here("figures", "roots_total_gm2.png"), plot = roots_total_gm2, width = 10, height = 8)

```

```{r}
#|label: column-graph-ANPP-to-rootbiomass-by-trt-and-site

root_shoot_plot <- ANPPandRoot_df |>
  ggplot(aes(x = SiteOrder, y = root_to_ANPP, fill = TreatmentOrder)) + 
  geom_boxplot(position = position_dodge(0.8)) +  
  scale_fill_manual(name = "", values = c("#de2d26", "#bdbdbd", "#08519c")) +
  scale_y_continuous(limits = c(0, 26), expand = c(0, 0)) +
labs(
   x = "Site", 
   y = expression("Root Mass:ANPP")) +
  scale_x_discrete(labels = site_labels) +
  theme_grey(base_size = 18) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.ticks.x = element_blank(),
    legend.position = c(0.3, 0.7),
    panel.background = element_blank(),
    panel.border = element_rect(color = "black", fill = NA)
  ) 
root_shoot_plot

ggsave(here("figures", "root_to_ANPP.png"), plot = root_shoot_plot, width = 10, height = 8)
```

## ANOVAs

```{r}
#| label: Canopy root biomass ANOVA

# Define output file
output_file <- here("results", "ANOVA_Root_mass_Can_gm2.txt")
sink(output_file)

cat("=== Normality Check and Transformation ===\n\n")

# --- Raw normality ---
cat("Shapiro-Wilk test (raw data):\n")
shapiro_raw <- shapiro.test(ANPPandRoot_df$Root_mass_Can_gm2)
print(shapiro_raw)

# Apply bestNormalize if not normal
if (shapiro_raw$p.value < 0.05) {
  cat("\nData not normal (p < 0.05). Applying bestNormalize...\n")
  BN_RootCan_gm2 <- bestNormalize(ANPPandRoot_df$Root_mass_Can_gm2)
  print(BN_RootCan_gm2)
  
  ANPPandRoot_df$Root_mass_Can_gm2_norm <- predict(BN_RootCan_gm2)
  
  cat("\nShapiro-Wilk test (transformed data):\n")
  shapiro_trans <- shapiro.test(ANPPandRoot_df$Root_mass_Can_gm2_norm)
  print(shapiro_trans)
  
  used_var <- "Root_mass_Can_gm2_norm"
  transform_used <- class(BN_RootCan_gm2$chosen_transform)[1]
  
} else {
  cat("\nData are already normal (p >= 0.05). No transformation applied.\n")
  ANPPandRoot_df$Root_mass_Can_gm2_norm <- ANPPandRoot_df$Root_mass_Can_gm2
  used_var <- "Root_mass_Can_gm2"
  transform_used <- "None"
}

cat("\nTransformation method used:", transform_used, "\n")

# --- QQ plots ---
par(mfrow = c(1, 2))
qqnorm(ANPPandRoot_df$Root_mass_Can_gm2, main = "Raw Data")
qqline(ANPPandRoot_df$Root_mass_Can_gm2)
qqnorm(ANPPandRoot_df$Root_mass_Can_gm2_norm, main = "Transformed Data")
qqline(ANPPandRoot_df$Root_mass_Can_gm2_norm)
par(mfrow = c(1, 1))

# --- ANOVA ---
cat("\n\n=== Two-Way ANOVA ===\n\n")
anova_formula <- as.formula(paste(used_var, "~ Site * Treatment"))
Root_Can_gm2_anova <- aov(anova_formula, data = ANPPandRoot_df)
anova_summary <- summary(Root_Can_gm2_anova)
print(anova_summary)

# --- Tukey’s HSD ---
cat("\n\n=== Pairwise Comparisons (Tukey's HSD) ===\n\n")
Can_pairwise <- TukeyHSD(Root_Can_gm2_anova)
print(Can_pairwise)

sink()  # close the file
```

```{r}
#| label: IC root biomass ANOVA

# Define output file
output_file <- here("results", "ANOVA_Root_mass_IC_gm2.txt")
sink(output_file)

cat("=== Normality Check and Transformation ===\n\n")

# --- Raw normality ---
cat("Shapiro-Wilk test (raw data):\n")
shapiro_raw <- shapiro.test(ANPPandRoot_df$Root_mass_IC_gm2)
print(shapiro_raw)

# Apply bestNormalize if not normal
if (shapiro_raw$p.value < 0.05) {
  cat("\nData not normal (p < 0.05). Applying bestNormalize...\n")
  BN_RootIC_gm2 <- bestNormalize(ANPPandRoot_df$Root_mass_IC_gm2)
  print(BN_RootIC_gm2)
  
  ANPPandRoot_df$Root_mass_IC_gm2_norm <- predict(BN_RootIC_gm2)
  
  cat("\nShapiro-Wilk test (transformed data):\n")
  shapiro_trans <- shapiro.test(ANPPandRoot_df$Root_mass_IC_gm2_norm)
  print(shapiro_trans)
  
  used_var <- "Root_mass_IC_gm2_norm"
  transform_used <- class(BN_RootIC_gm2$chosen_transform)[1]
  
} else {
  cat("\nData are already normal (p >= 0.05). No transformation applied.\n")
  ANPPandRoot_df$Root_mass_IC_gm2_norm <- ANPPandRoot_df$Root_mass_IC_gm2
  used_var <- "Root_mass_IC_gm2"
  transform_used <- "None"
}

cat("\nTransformation method used:", transform_used, "\n")

# --- QQ plots ---
par(mfrow = c(1, 2))
qqnorm(ANPPandRoot_df$Root_mass_IC_gm2, main = "Raw Data")
qqline(ANPPandRoot_df$Root_mass_IC_gm2)
qqnorm(ANPPandRoot_df$Root_mass_IC_gm2_norm, main = "Transformed Data")
qqline(ANPPandRoot_df$Root_mass_IC_gm2_norm)
par(mfrow = c(1, 1))

# --- ANOVA ---
cat("\n\n=== Two-Way ANOVA ===\n\n")
anova_formula <- as.formula(paste(used_var, "~ Site * Treatment"))
Root_IC_gm2_anova <- aov(anova_formula, data = ANPPandRoot_df)
anova_summary <- summary(Root_IC_gm2_anova)
print(anova_summary)

# --- Tukey’s HSD ---
cat("\n\n=== Pairwise Comparisons (Tukey's HSD) ===\n\n")
IC_pairwise <- TukeyHSD(Root_IC_gm2_anova)
print(IC_pairwise)

sink()  # close the file

```

### 

using area-weighted values for whole plot root mass (g m-2)

```{r}
#| label: total root biomass ANOVA

# Define output file
output_file <- here("results", "ANOVA_Root_mass_total_gm2.txt")
sink(output_file)

cat("=== Normality Check and Transformation ===\n\n")

# --- Raw normality ---
cat("Shapiro-Wilk test (raw data):\n")
shapiro_raw <- shapiro.test(ANPPandRoot_df$Root_mass_total_gm2)
print(shapiro_raw)

# Apply bestNormalize if not normal
if (shapiro_raw$p.value < 0.05) {
  cat("\nData not normal (p < 0.05). Applying bestNormalize...\n")
  BN_Roottotal_gm2 <- bestNormalize(ANPPandRoot_df$Root_mass_total_gm2)
  print(BN_Roottotal_gm2)
  
  ANPPandRoot_df$Root_mass_total_gm2_norm <- predict(BN_Roottotal_gm2)
  
  cat("\nShapiro-Wilk test (transformed data):\n")
  shapiro_trans <- shapiro.test(ANPPandRoot_df$Root_mass_total_gm2_norm)
  print(shapiro_trans)
  
  used_var <- "Root_mass_total_gm2_norm"
  transform_used <- class(BN_Roottotal_gm2$chosen_transform)[1]
  
} else {
  cat("\nData are already normal (p >= 0.05). No transformation applied.\n")
  ANPPandRoot_df$Root_mass_total_gm2_norm <- ANPPandRoot_df$Root_mass_total_gm2
  used_var <- "Root_mass_total_gm2"
  transform_used <- "None"
}

cat("\nTransformation method used:", transform_used, "\n")

# --- QQ plots ---
par(mfrow = c(1, 2))
qqnorm(ANPPandRoot_df$Root_mass_total_gm2, main = "Raw Data")
qqline(ANPPandRoot_df$Root_mass_total_gm2)
qqnorm(ANPPandRoot_df$Root_mass_total_gm2_norm, main = "Transformed Data")
qqline(ANPPandRoot_df$Root_mass_total_gm2_norm)
par(mfrow = c(1, 1))

# --- ANOVA ---
cat("\n\n=== Two-Way ANOVA ===\n\n")
anova_formula <- as.formula(paste(used_var, "~ Site * Treatment"))
Root_total_gm2_anova <- aov(anova_formula, data = ANPPandRoot_df)
anova_summary <- summary(Root_total_gm2_anova)
print(anova_summary)

# --- Tukey’s HSD ---
cat("\n\n=== Pairwise Comparisons (Tukey's HSD) ===\n\n")
total_pairwise <- TukeyHSD(Root_total_gm2_anova)
print(total_pairwise)

sink()  # close the file
```

### ANPP

```{r}
# check normality and transform as needed
qqnorm(ANPPandRoot_df$plot_mean_ANPP) 
shapiro.test(ANPPandRoot_df$plot_mean_ANPP) # P < 0.05 - reject hypothesis of normality
BN_ANPP <- bestNormalize(ANPPandRoot_df$plot_mean_ANPP)
BN_ANPP # show options for normalizing
ANPPandRoot_df$plot_mean_ANPP_norm <- predict(BN_ANPP)
qqnorm(ANPPandRoot_df$plot_mean_ANPP_norm) # a little better
qqline(ANPPandRoot_df$plot_mean_ANPP_norm) # add line to plot
shapiro.test(ANPPandRoot_df$plot_mean_ANPP_norm) # P > 0.05 - ok

# 2-way ANOVA model with transformed data
ANPP_anova <- aov(plot_mean_ANPP_norm ~ Site * Treatment, data=ANPPandRoot_df)
summary(ANPP_anova)

# Perform pairwise comparisons using Tukey's HSD test
ANPP_pairwise <- TukeyHSD(ANPP_anova)
print(ANPP_pairwise)
plot(ANPP_pairwise)


### START HERE -- modify the code below for ANPP output. Do this also for Root:ANPP.

#| label: IC root biomass ANOVA

# Define output file
output_file <- here("results", "ANOVA_Root_mass_IC_gm2.txt")
sink(output_file)

cat("=== Normality Check and Transformation ===\n\n")

# --- Raw normality ---
cat("Shapiro-Wilk test (raw data):\n")
shapiro_raw <- shapiro.test(ANPPandRoot_df$Root_mass_IC_gm2)
print(shapiro_raw)

# Apply bestNormalize if not normal
if (shapiro_raw$p.value < 0.05) {
  cat("\nData not normal (p < 0.05). Applying bestNormalize...\n")
  BN_RootIC_gm2 <- bestNormalize(ANPPandRoot_df$Root_mass_IC_gm2)
  print(BN_RootIC_gm2)
  
  ANPPandRoot_df$Root_mass_IC_gm2_norm <- predict(BN_RootIC_gm2)
  
  cat("\nShapiro-Wilk test (transformed data):\n")
  shapiro_trans <- shapiro.test(ANPPandRoot_df$Root_mass_IC_gm2_norm)
  print(shapiro_trans)
  
  used_var <- "Root_mass_IC_gm2_norm"
  transform_used <- class(BN_RootIC_gm2$chosen_transform)[1]
  
} else {
  cat("\nData are already normal (p >= 0.05). No transformation applied.\n")
  ANPPandRoot_df$Root_mass_IC_gm2_norm <- ANPPandRoot_df$Root_mass_IC_gm2
  used_var <- "Root_mass_IC_gm2"
  transform_used <- "None"
}

cat("\nTransformation method used:", transform_used, "\n")

# --- QQ plots ---
par(mfrow = c(1, 2))
qqnorm(ANPPandRoot_df$Root_mass_IC_gm2, main = "Raw Data")
qqline(ANPPandRoot_df$Root_mass_IC_gm2)
qqnorm(ANPPandRoot_df$Root_mass_IC_gm2_norm, main = "Transformed Data")
qqline(ANPPandRoot_df$Root_mass_IC_gm2_norm)
par(mfrow = c(1, 1))

# --- ANOVA ---
cat("\n\n=== Two-Way ANOVA ===\n\n")
anova_formula <- as.formula(paste(used_var, "~ Site * Treatment"))
Root_IC_gm2_anova <- aov(anova_formula, data = ANPPandRoot_df)
anova_summary <- summary(Root_IC_gm2_anova)
print(anova_summary)

# --- Tukey’s HSD ---
cat("\n\n=== Pairwise Comparisons (Tukey's HSD) ===\n\n")
IC_pairwise <- TukeyHSD(Root_IC_gm2_anova)
print(IC_pairwise)

sink()  # close the file

```

### Root:ANPP

```{r}
# check normality and transform as needed
qqnorm(ANPPandRoot_df$root_to_ANPP) 
shapiro.test(ANPPandRoot_df$root_to_ANPP) # P < 0.05 - reject hypothesis of normality
BN_root_to_ANPP <- bestNormalize(ANPPandRoot_df$root_to_ANPP)
BN_root_to_ANPP # show options for normalizing
ANPPandRoot_df$root_to_ANPP_norm <- predict(BN_root_to_ANPP)
qqnorm(ANPPandRoot_df$root_to_ANPP_norm) # a little better
qqline(ANPPandRoot_df$root_to_ANPP_norm) # add line to plot
shapiro.test(ANPPandRoot_df$root_to_ANPP_norm) # P > 0.05 - ok

# 2-way ANOVA model with transformed data
root_to_ANPP_anova <- aov(root_to_ANPP_norm ~ Site * Treatment, data=ANPPandRoot_df)
summary(root_to_ANPP_anova)

# Perform pairwise comparisons using Tukey's HSD test
root_to_ANPP_pairwise <- TukeyHSD(root_to_ANPP_anova)
print(root_to_ANPP_pairwise)
plot(root_to_ANPP_pairwise)

```

## Combined Plots

```{r}
library(patchwork)

# Label plots directly using patchwork tag system
library(patchwork)

# Common theme for alignment
base_theme <- theme(
  axis.title.y = element_text(size = 16),
  plot.margin = margin(t = 5.5, r = 5.5, b = 5.5, l = 14),
  panel.border = element_rect(color = "black", fill = NA)
)

# Apply base theme
p_top <- ANPP +
  scale_x_discrete(labels = site_labels) +
  theme(
    axis.title.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_line(),
    legend.position = c(0.2, 0.7),
  ) + base_theme

p_middle <- roots_total_gm2 +
  scale_x_discrete(labels = site_labels) +
  theme(
    axis.title.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_line(),
    legend.position = "none"
  ) + base_theme

p_bottom <- root_shoot_plot +
  scale_x_discrete(labels = site_labels) +
  theme(
    axis.title.x = element_text(size = 16),
    axis.ticks.x = element_line(),
    legend.position = "none"
  ) + base_theme

# Combine plots with letter tags in upper left, with "." after letters
combined_plot <- (p_top / p_middle / p_bottom) +
  plot_annotation(tag_levels = 'A', tag_suffix = ".", tag_prefix = "") &
  theme(
    plot.tag = element_text(size = 16, face = "plain", hjust = 0, vjust = 1),
    plot.tag.position = c(0.1, 0.95)  # upper-left corner
  )

combined_plot

ggsave(
  filename = here("figures", "ANPP_rootbiomass.png"),  
  plot = combined_plot,
  width = 12,       # width in inches
  height = 8,       # height in inches
  dpi = 300         # high-resolution PNG
)

```
