---
title: "GRaMPS_CN"
author: "Heather Throop"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

### Overview

This is a compilation, cleaning, and analysis of soil carbon and nitrogen data from Brittney Monus's MS research in the GRaMPS project.

Code was written by Heather Throop, streamlining and updating the original analyses that Brittney did for her MS.

Updated:\
2025-07-25 Heather Throop\
2025-10-01 Heather Throop - added N ANOVA summary output 2025-10-03 Heather Throop - added graphs for 5-10 cm depths

## SetUp

```{r}
#| label: load packages
library(tidyverse)
library(car)       # For Levene's test
library(emmeans)   # For post-hoc comparisons
library(scales)
library(here)
library(bestNormalize)
library(car)
library(broom)
```

## Ingest and Clean L1 Data

This chunk can be skipped - start with ingesting L2 data (below)

```{r}
#| label: prepare data

## read in L1 data - this is the daily corrected CN data for all runs. 
CN = read.csv(here("data","Gramps_CN_L1.csv"))

## pull identifying information from the string in the ID column

# Define mapping vectors
site_map <- c("BP" = "Desert Scrub",
              "ANT" = "Desert Grassland",
              "BC" = "Juniper Savanna",
              "ARB" = "Ponderosa Pine Meadow",
              "CC" = "Mixed Conifer Meadow")
treatment_map <- c("exl" = "Exclusion",
                   "ctrl" = "Ambient",
                   "add" = "Addition")

# Clean and parse the ID column
CN_clean <- CN |>
  mutate(
    # Extract components using regular expressions
    Site_code = str_extract(ID, "^[A-Z]{2,3}"),
    Treatment_code = str_extract(ID, "(?<=^[A-Z]{2,3})[a-z]{3,4}"),
    Plot = str_extract(ID, "(?<=P)\\d{1,2}"),
    Microsite = str_extract(ID, "(CAN|IC)"),
    Depth_raw = str_extract(ID, "(05|510)"),
    
    # Clean extracted values
    Site = dplyr::recode(Site_code, !!!site_map),
    Treatment = dplyr::recode(Treatment_code, !!!treatment_map),
    Depth = str_remove(Depth_raw, "_") |> 
            str_replace("05", "0.5") |> 
            str_replace("510", "5.10") |> 
            as.factor(),
    Depth = as.factor(Depth)) |>
    dplyr::select(-Site_code, -Treatment_code, -Depth_raw)  # drop intermediate columns

# Preview cleaned data
glimpse(CN_clean)

# rename and order sites, treatments, and microsites
CN_clean <- CN_clean |>
      mutate(Site = factor(Site, levels = 
                               c("Desert Scrub","Desert Grassland","Juniper Savanna",
                                       "Ponderosa Pine Meadow","Mixed Conifer Meadow")),
             Treatment = factor(Treatment, levels = 
                                  c("Exclusion","Ambient","Addition")),
             Microsite = factor(Microsite, 
                                   levels=c("IC", "CAN"), 
                                   labels=c("Intercanopy", "Canopy")) )

### reduce to plot means
# calculate mean and SE for each Site x Treatment x Microsite x Depth x Plot combination. In most cases there was only one analytical replicate per plot, but this accounts for cases where there were analytical reps.
CN_plot.microsite_means <- CN_clean |>
  group_by(Site, Treatment, Microsite, Depth, Plot) |>
  summarise(
    N = n(),
    mean_C = mean(pctC, na.rm = TRUE),
    mean_N = mean(pctN, na.rm = TRUE),
    .groups = 'drop'
  )

# save clean L2 data ready for analysis
write.csv(CN_plot.microsite_means, 
          here("data","calculated_outputs","CN_L2.csv"), row.names = FALSE)

```

## Microsite Means

Start here to pull in cleaned L2 data

```{r}
#| label: load L2 data & split by depth

## read in L2 data - one value for each site x treatment x plot x microsite x depth combination
CN_L2 = read.csv(here("data","calculated_outputs","CN_L2.csv"))

# Ensure factors are properly coded
CN_L2 <- CN_L2 |>
  mutate(across(c(Site, Treatment, Microsite, Depth), as.factor))

```

### 0-5 cm depth

#### carbon

```{r}
#| label: Check ANOVA assumptions - pctC 

# Filter and prepare data
CN_0.5 <- CN_L2 |>
  filter(Depth == "0.5") |>
  mutate(Group = interaction(Site, Treatment, Microsite))

# Step 1: Check initial normality
hist(CN_0.5$mean_C, main = "Histogram of mean_C (0–5 cm)", xlab = "mean_C")
qqnorm(CN_0.5$mean_C); qqline(CN_0.5$mean_C)

shapiro_raw <- shapiro.test(CN_0.5$mean_C)
cat("Shapiro-Wilk p-value (raw):", shapiro_raw$p.value, "\n")

# Step 2: Apply bestNormalize if needed
if (shapiro_raw$p.value < 0.05) {
  bn_obj <- bestNormalize(CN_0.5$mean_C, allow_orderNorm = FALSE)
  CN_0.5 <- CN_0.5 %>%
    mutate(C_trans = predict(bn_obj, newdata = mean_C))
  used_var <- "C_trans"
  transform_used <- class(bn_obj$chosen_transform)[1]
  
  # Check transformed normality
  hist(CN_0.5$C_trans, main = "Histogram of Transformed C", xlab = "C_trans")
  qqnorm(CN_0.5$C_trans); qqline(CN_0.5$C_trans)
  
  shapiro_trans <- shapiro.test(CN_0.5$C_trans)
  cat("Shapiro-Wilk p-value (transformed):", shapiro_trans$p.value, "\n")
} else {
  used_var <- "mean_C"
  transform_used <- "None"
}

# Step 3: Levene's Test
levene_result <- leveneTest(reformulate("Group", used_var), data = CN_0.5)

# Step 4: Summary Table
summary_table <- tibble(
  variable = used_var,
  transformed = transform_used,
  shapiro_p_raw = shapiro_raw$p.value,
  shapiro_p_trans = if (exists("shapiro_trans")) shapiro_trans$p.value else NA,
  levene_p = levene_result$`Pr(>F)`[1]
)
print(summary_table)

```

```{r}
#| label: three-way ANOVA

# Three-way ANOVA model
# Step 5: Run 3-way ANOVA
anova_model <- aov(reformulate(c("Site", "Treatment", "Microsite",
                                  "Site:Treatment",
                                  "Site:Microsite",
                                  "Treatment:Microsite",
                                  "Site:Treatment:Microsite"),
                                used_var),
                   data = CN_0.5)
anova_summary <- summary(anova_model)
print(anova_summary)

group_summary <- CN_0.5 |>
  group_by(Site, Treatment, Microsite) |>
  summarise(
    mean_val = mean(C_trans, na.rm = TRUE),
    se_val   = sd(C_trans, na.rm = TRUE) / sqrt(sum(!is.na(C_trans))),
    .groups = "drop"
  )

group_summary1 <- CN_0.5 |>
  group_by(Site, Treatment, Microsite) |>
  summarise(
    mean_val = mean(mean_C, na.rm = TRUE),
    se_val   = sd(mean_C, na.rm = TRUE) / sqrt(sum(!is.na(mean_C))),
    .groups = "drop"
  )

# --- Site main effect ---
site_summary <- CN_0.5 |>
  group_by(Site) |>
  summarise(
    mean_val = mean(mean_C, na.rm = TRUE),
    se_val   = sd(mean_C, na.rm = TRUE) / sqrt(sum(!is.na(mean_C))),
    .groups = "drop"
  )

# --- Treatment main effect ---
treat_summary <- CN_0.5 |>
  group_by(Treatment) |>
  summarise(
    mean_val = mean(mean_C, na.rm = TRUE),
    se_val   = sd(mean_C, na.rm = TRUE) / sqrt(sum(!is.na(mean_C))),
    .groups = "drop"
  )

# --- Microsite main effect ---
micro_summary <- CN_0.5 |>
  group_by(Microsite) |>
  summarise(
    mean_val = mean(mean_C, na.rm = TRUE),
    se_val   = sd(mean_C, na.rm = TRUE) / sqrt(sum(!is.na(mean_C))),
    .groups = "drop"
  )

main_effects_summary <- bind_rows(
  site_summary   %>% mutate(effect = "Site"),
  treat_summary  %>% mutate(effect = "Treatment"),
  micro_summary  %>% mutate(effect = "Microsite")
)


# Step 6: Save output to a text file
sink(here::here("results", "anova_summary_pctC_0.5.txt"))
cat("=== Transformation Summary ===\n")
print(summary_table)
cat("\n=== ANOVA Summary ===\n")
print(anova_summary)
cat("\n=== Group Means and Standard Errors for trans_C and pctC ===\n")
print(group_summary)
print(group_summary1)
print(main_effects_summary)
sink()
```

```{r}
#| label: boxplot for 0-5 cm %C

# Microsite labels get screwed up in running this code multiple times - rerun from start 

# put the microsite, site, and treatment in order
CN_0.5 <- CN_0.5 |>
      mutate(
        Microsite = factor(Microsite, levels = c("Intercanopy", "Canopy"), 
                           labels = c("IC", "Canopy")),
        Treatment = factor(Treatment, levels = c("Exclusion","Ambient","Addition")),
        Site = factor(Site, levels = c("Desert Scrub","Desert Grassland",
                                       "Juniper Savanna", "Ponderosa Pine Meadow",
                                       "Mixed Conifer Meadow"))
        )

panel_labels <- data.frame(
  Site = unique(CN_0.5$Site),
  x = 0.6,  # Nudges left of the first x category
  y = 5.9,
  label = c("A.", "B.", "C.", "D.", "E.")
)

p_pctC_0.5 <- CN_0.5 |> 
  ggplot(aes(x = Microsite, y = mean_C, fill = Treatment)) + 
  geom_boxplot() +
  facet_wrap(~ Site, nrow = 1, labeller = labeller(Site = label_wrap_gen(width = 15))) +
    geom_text(
    data = panel_labels,
    inherit.aes = FALSE,
    aes(x = x, y = y, label = label),
    hjust = 0, vjust = 1,
    size = 5) +
  ylab("Soil Organic Carbon (%)") +
  xlab("Microsite") +
  theme_grey(base_size = 18) +
  scale_fill_manual(name = "", values = c("#de2d26", "#bdbdbd", "#08519c")) +
  theme_grey(base_size = 18) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "white", color = NA),
    axis.text.x = element_text(size = 14),
    axis.text = element_text(size = 15),
    panel.border = element_rect(color = "black", fill = NA),
    legend.position = c(0.001, 0.88),
    legend.justification = c("left", "top")
  ) +
   scale_y_continuous(limits = c(0, 6), breaks = c(0, 1.5, 3.0, 4.5, 6.0), 
                     expand = c(0, 0)) +
  guides(color = guide_legend(title = NULL))
p_pctC_0.5

ggsave(
  filename = here("figures", "pctC_0.5cm.png"),  
  plot = p_pctC_0.5,
  width = 12,       # width in inches
  height = 6,       # height in inches
  dpi = 300         # high-resolution PNG
)
```

#### nitrogen

```{r}
#| label: Check ANOVA assumptions - pctN 

# Filter and prepare data
CN_0.5 <- CN_L2 |>
  filter(Depth == "0.5") |>
  mutate(Group = interaction(Site, Treatment, Microsite))

# Step 1: Check initial normality
hist(CN_0.5$mean_N, main = "Histogram of mean_N (0–5 cm)", xlab = "mean_N")
qqnorm(CN_0.5$mean_N); qqline(CN_0.5$mean_N)

shapiro_raw <- shapiro.test(CN_0.5$mean_N)
cat("Shapiro-Wilk p-value (raw):", shapiro_raw$p.value, "\n")

# Step 2: Apply bestNormalize if needed
if (shapiro_raw$p.value < 0.05) {
  bn_obj <- bestNormalize(CN_0.5$mean_N, allow_orderNorm = FALSE)
  CN_0.5 <- CN_0.5 %>%
    mutate(N_trans = predict(bn_obj, newdata = mean_N))
  used_var <- "N_trans"
  transform_used <- class(bn_obj$chosen_transform)[1]
  
  # Check transformed normality
  hist(CN_0.5$N_trans, main = "Histogram of Transformed N", xlab = "N_trans")
  qqnorm(CN_0.5$N_trans); qqline(CN_0.5$N_trans)
  
  shapiro_trans <- shapiro.test(CN_0.5$N_trans)
  cat("Shapiro-Wilk p-value (transformed):", shapiro_trans$p.value, "\n")
} else {
  used_var <- "mean_N"
  transform_used <- "None"
}

# Step 3: Levene's Test
levene_result <- leveneTest(reformulate("Group", used_var), data = CN_0.5)

# Step 4: Summary Table
summary_table <- tibble(
  variable = used_var,
  transformed = transform_used,
  shapiro_p_raw = shapiro_raw$p.value,
  shapiro_p_trans = if (exists("shapiro_trans")) shapiro_trans$p.value else NA,
  levene_p = levene_result$`Pr(>F)`[1]
)
print(summary_table)

```

```{r}
#| label: three-way ANOVA

# Three-way ANOVA model
# Step 5: Run 3-way ANOVA
anova_model <- aov(reformulate(c("Site", "Treatment", "Microsite",
                                  "Site:Treatment",
                                  "Site:Microsite",
                                  "Treatment:Microsite",
                                  "Site:Treatment:Microsite"),
                                used_var),
                   data = CN_0.5)
anova_summary <- summary(anova_model)
print(anova_summary)

group_summary <- CN_0.5 |>
  group_by(Site, Treatment, Microsite) |>
  summarise(
    mean_val = mean(N_trans, na.rm = TRUE),
    se_val   = sd(N_trans, na.rm = TRUE) / sqrt(sum(!is.na(N_trans))),
    .groups = "drop"
  )

group_summary1 <- CN_0.5 |>
  group_by(Site, Treatment, Microsite) |>
  summarise(
    mean_val = mean(mean_N, na.rm = TRUE),
    se_val   = sd(mean_N, na.rm = TRUE) / sqrt(sum(!is.na(mean_N))),
    .groups = "drop"
  )

# --- Site main effect ---
site_summary <- CN_0.5 |>
  group_by(Site) |>
  summarise(
    mean_val = mean(mean_N, na.rm = TRUE),
    se_val   = sd(mean_N, na.rm = TRUE) / sqrt(sum(!is.na(mean_N))),
    .groups = "drop"
  )

# --- Treatment main effect ---
treat_summary <- CN_0.5 |>
  group_by(Treatment) |>
  summarise(
    mean_val = mean(mean_N, na.rm = TRUE),
    se_val   = sd(mean_N, na.rm = TRUE) / sqrt(sum(!is.na(mean_N))),
    .groups = "drop"
  )

# --- Microsite main effect ---
micro_summary <- CN_0.5 |>
  group_by(Microsite) |>
  summarise(
    mean_val = mean(mean_N, na.rm = TRUE),
    se_val   = sd(mean_N, na.rm = TRUE) / sqrt(sum(!is.na(mean_N))),
    .groups = "drop"
  )

main_effects_summary <- bind_rows(
  site_summary   %>% mutate(effect = "Site"),
  treat_summary  %>% mutate(effect = "Treatment"),
  micro_summary  %>% mutate(effect = "Microsite")
)

# Step 6: Save output to a text file
sink(here::here("results", "anova_summary_pctN_0.5.txt"))
cat("=== Transformation Summary ===\n")
print(summary_table)
cat("\n=== ANOVA Summary for pct N 0-5 cm data  ===\n")
print(anova_summary)
cat("\n=== Group Means and Standard Errors for trans_N and pctN ===\n")
print(group_summary)
print(group_summary1)
print(main_effects_summary)
sink()
```

```{r}
#| label: boxplot for 0-5 cm %N

# Microsite labels get screwed up in running this code multiple times - rerun from start 

# put the microsite, site, and treatment in order
CN_0.5 <- CN_0.5 |>
      mutate(
        Microsite = factor(Microsite, levels = c("Intercanopy", "Canopy"), 
                           labels = c("IC", "Canopy")),
        Treatment = factor(Treatment, levels = c("Exclusion","Ambient","Addition")),
        Site = factor(Site, levels = c("Desert Scrub","Desert Grassland",
                                       "Juniper Savanna", "Ponderosa Pine Meadow",
                                       "Mixed Conifer Meadow"))
        )

panel_labels <- data.frame(
  Site = unique(CN_0.5$Site),
  x = 0.6,  # Nudges left of the first x category
  y = 0.49,
  label = c("F.", "G.", "H.", "I.", "J.")
)

p_pctN_0.5 <- CN_0.5 |> 
  ggplot(aes(x = Microsite, y = mean_N, fill = Treatment)) + 
  geom_boxplot() +
  facet_wrap(~ Site, nrow = 1, labeller = labeller(Site = label_wrap_gen(width = 15))) +
  geom_text(
    data = panel_labels,
    inherit.aes = FALSE,
    aes(x = x, y = y, label = label),
    hjust = 0, vjust = 1,
    size = 5) +
  scale_y_continuous(limits = c(0, 0.5), breaks = c(0, 0.1, 0.2, 0.3, 0.4, 0.5), 
                     expand = c(0, 0)) +
  ylab("Total Nitrogen (%)") +
  xlab("Microsite") +
  theme_grey(base_size = 18) +
  scale_fill_manual(name = "", values = c("#de2d26", "#bdbdbd", "#08519c")) +
  theme_grey(base_size = 18) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "white", color = NA),
    axis.text.x = element_text(size = 14),
    axis.text = element_text(size = 15),
    panel.border = element_rect(color = "black", fill = NA),
    legend.position = "none") +
  guides(color = guide_legend(title = NULL))
p_pctN_0.5

ggsave(
  filename = here("figures", "pctN_0.5cm.png"),  
  plot = p_pctN_0.5,
  width = 12,       # width in inches
  height = 6,       # height in inches
  dpi = 300         # high-resolution PNG
)
```

### 

### 5-10 cm depth

#### carbon

```{r}
#| label: Check ANOVA assumptions

# Filter and prepare data
CN_5.10 <- CN_L2 |>
  filter(Depth == "5.1") |>
  mutate(Group = interaction(Site, Treatment, Microsite))

# Step 1: Check initial normality
hist(CN_5.10$mean_C, main = "Histogram of mean_C (5-10 cm)", xlab = "mean_C")
qqnorm(CN_5.10$mean_C); qqline(CN_5.10$mean_C)

shapiro_raw <- shapiro.test(CN_5.10$mean_C)
cat("Shapiro-Wilk p-value (raw):", shapiro_raw$p.value, "\n")

# Step 2: Apply bestNormalize if needed
if (shapiro_raw$p.value < 0.05) {
  bn_obj <- bestNormalize(CN_5.10$mean_C, allow_orderNorm = FALSE)
  CN_5.10 <- CN_5.10 %>%
    mutate(C_trans = predict(bn_obj, newdata = mean_C))
  used_var <- "C_trans"
  transform_used <- class(bn_obj$chosen_transform)[1]
  
  # Check transformed normality
  hist(CN_5.10$C_trans, main = "Histogram of Transformed C", xlab = "C_trans")
  qqnorm(CN_5.10$C_trans); qqline(CN_5.10$C_trans)
  
  shapiro_trans <- shapiro.test(CN_5.10$C_trans)
  cat("Shapiro-Wilk p-value (transformed):", shapiro_trans$p.value, "\n")
} else {
  used_var <- "mean_C"
  transform_used <- "None"
}

# Step 3: Levene's Test
levene_result <- leveneTest(reformulate("Group", used_var), data = CN_5.10)

# Step 4: Summary Table
summary_table <- tibble(
  variable = used_var,
  transformed = transform_used,
  shapiro_p_raw = shapiro_raw$p.value,
  shapiro_p_trans = if (exists("shapiro_trans")) shapiro_trans$p.value else NA,
  levene_p = levene_result$`Pr(>F)`[1]
)
print(summary_table)

```

```{r}
#| label: three-way ANOVA

# Three-way ANOVA model
# Step 5: Run 3-way ANOVA
anova_model <- aov(reformulate(c("Site", "Treatment", "Microsite",
                                  "Site:Treatment",
                                  "Site:Microsite",
                                  "Treatment:Microsite",
                                  "Site:Treatment:Microsite"),
                                used_var),
                   data = CN_5.10)
anova_summary <- summary(anova_model)
print(anova_summary)

group_summary <- CN_5.10 |>
  group_by(Site, Treatment, Microsite) |>
  summarise(
    mean_val = mean(C_trans, na.rm = TRUE),
    se_val   = sd(C_trans, na.rm = TRUE) / sqrt(sum(!is.na(C_trans))),
    .groups = "drop"
  )

group_summary1 <- CN_5.10 |>
  group_by(Site, Treatment, Microsite) |>
  summarise(
    mean_val = mean(mean_C, na.rm = TRUE),
    se_val   = sd(mean_C, na.rm = TRUE) / sqrt(sum(!is.na(mean_C))),
    .groups = "drop"
  )

# --- Site main effect ---
site_summary <- CN_5.10 |>
  group_by(Site) |>
  summarise(
    mean_val = mean(mean_C, na.rm = TRUE),
    se_val   = sd(mean_C, na.rm = TRUE) / sqrt(sum(!is.na(mean_C))),
    .groups = "drop"
  )

# --- Treatment main effect ---
treat_summary <- CN_5.10 |>
  group_by(Treatment) |>
  summarise(
    mean_val = mean(mean_C, na.rm = TRUE),
    se_val   = sd(mean_C, na.rm = TRUE) / sqrt(sum(!is.na(mean_C))),
    .groups = "drop"
  )

# --- Microsite main effect ---
micro_summary <- CN_5.10 |>
  group_by(Microsite) |>
  summarise(
    mean_val = mean(mean_C, na.rm = TRUE),
    se_val   = sd(mean_C, na.rm = TRUE) / sqrt(sum(!is.na(mean_C))),
    .groups = "drop"
  )

main_effects_summary <- bind_rows(
  site_summary   %>% mutate(effect = "Site"),
  treat_summary  %>% mutate(effect = "Treatment"),
  micro_summary  %>% mutate(effect = "Microsite")
)

# Step 6: Save output to a text file
sink(here::here("results", "anova_summary_pctC_5.10.txt"))
cat("=== Transformation Summary ===\n")
print(summary_table)
cat("\n=== ANOVA Summary for %C 5-10 cm depth===\n")
print(anova_summary)
cat("\n=== Group Means and Standard Errors for trans_C and pctC ===\n")
print(group_summary)
print(group_summary1)
print(main_effects_summary)
sink()
```

```{r}
#| label: boxplot for 5-10 cm %C

# Microsite labels get screwed up in running this code multiple times - rerun from start 

# put the microsite, site, and treatment in order
CN_5.10 <- CN_5.10 |>
      mutate(
        Microsite = factor(Microsite, levels = c("Intercanopy", "Canopy"), 
                           labels = c("IC", "Canopy")),
        Treatment = factor(Treatment, levels = c("Exclusion","Ambient","Addition")),
        Site = factor(Site, levels = c("Desert Scrub","Desert Grassland",
                                       "Juniper Savanna", "Ponderosa Pine Meadow",
                                       "Mixed Conifer Meadow"))
        )

panel_labels <- data.frame(
  Site = unique(CN_5.10$Site),
  x = 0.6,  # Nudges left of the first x category
  y = 5.9,
  label = c("A.", "B.", "C.", "D.", "E.")
)

p_pctC_5.10 <- CN_5.10 |> 
  ggplot(aes(x = Microsite, y = mean_C, fill = Treatment)) + 
  geom_boxplot() +
  facet_wrap(~ Site, nrow = 1, labeller = labeller(Site = label_wrap_gen(width = 15))) +
    geom_text(
    data = panel_labels,
    inherit.aes = FALSE,
    aes(x = x, y = y, label = label),
    hjust = 0, vjust = 1,
    size = 5) +
  ylab("Soil Organic Carbon (%)") +
  xlab("Microsite") +
  theme_grey(base_size = 18) +
  scale_fill_manual(name = "", values = c("#de2d26", "#bdbdbd", "#08519c")) +
  theme_grey(base_size = 18) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "white", color = NA),
    axis.text.x = element_text(size = 14),
    axis.text = element_text(size = 15),
    panel.border = element_rect(color = "black", fill = NA),
    legend.position = c(0.001, 0.88),
    legend.justification = c("left", "top")
  ) +
   scale_y_continuous(limits = c(0, 6), breaks = c(0, 1.5, 3.0, 4.5, 6.0), 
                     expand = c(0, 0)) +
  guides(color = guide_legend(title = NULL))
p_pctC_5.10

ggsave(
  filename = here("figures", "pctC_5.10cm.png"),  
  plot = p_pctC_5.10,
  width = 12,       # width in inches
  height = 6,       # height in inches
  dpi = 300         # high-resolution PNG
)
```

### nitrogen

```{r}
#| label: Check ANOVA assumptions

# Filter and prepare data
CN_5.10 <- CN_L2 |>
  filter(Depth == "5.1") |>
  mutate(Group = interaction(Site, Treatment, Microsite))

# Step 1: Check initial normality
hist(CN_5.10$mean_N, main = "Histogram of mean_N (5-10 cm)", xlab = "mean_N")
qqnorm(CN_5.10$mean_N); qqline(CN_5.10$mean_N)

shapiro_raw <- shapiro.test(CN_5.10$mean_N)
cat("Shapiro-Wilk p-value (raw):", shapiro_raw$p.value, "\n")

# Step 2: Apply bestNormalize if needed
if (shapiro_raw$p.value < 0.05) {
  bn_obj <- bestNormalize(CN_5.10$mean_N, allow_orderNorm = FALSE)
  CN_5.10 <- CN_5.10 |>
    mutate(N_trans = predict(bn_obj, newdata = mean_N))
  used_var <- "N_trans"
  transform_used <- class(bn_obj$chosen_transform)[1]
  
  # Check transformed normality
  hist(CN_5.10$N_trans, main = "Histogram of Transformed N", xlab = "N_trans")
  qqnorm(CN_5.10$N_trans); qqline(CN_5.10$N_trans)
  
  shapiro_trans <- shapiro.test(CN_5.10$N_trans)
  cat("Shapiro-Wilk p-value (transformed):", shapiro_trans$p.value, "\n")
} else {
  used_var <- "mean_N"
  transform_used <- "None"
}

# Step 3: Levene's Test
levene_result <- leveneTest(reformulate("Group", used_var), data = CN_5.10)

# Step 4: Summary Table
summary_table <- tibble(
  variable = used_var,
  transformed = transform_used,
  shapiro_p_raw = shapiro_raw$p.value,
  shapiro_p_trans = if (exists("shapiro_trans")) shapiro_trans$p.value else NA,
  levene_p = levene_result$`Pr(>F)`[1]
)
print(summary_table)
```

```{r}
#| label: three-way ANOVA

# Three-way ANOVA model
# Step 5: Run 3-way ANOVA
anova_model <- aov(reformulate(c("Site", "Treatment", "Microsite",
                                  "Site:Treatment",
                                  "Site:Microsite",
                                  "Treatment:Microsite",
                                  "Site:Treatment:Microsite"),
                                used_var),
                   data = CN_5.10)
anova_summary <- summary(anova_model)
print(anova_summary)

group_summary <- CN_5.10 |>
  group_by(Site, Treatment, Microsite) |>
  summarise(
    mean_val = mean(N_trans, na.rm = TRUE),
    se_val   = sd(N_trans, na.rm = TRUE) / sqrt(sum(!is.na(N_trans))),
    .groups = "drop"
  )

group_summary1 <- CN_5.10 |>
  group_by(Site, Treatment, Microsite) |>
  summarise(
    mean_val = mean(mean_N, na.rm = TRUE),
    se_val   = sd(mean_N, na.rm = TRUE) / sqrt(sum(!is.na(mean_N))),
    .groups = "drop"
  )

# --- Site main effect ---
site_summary <- CN_5.10 |>
  group_by(Site) |>
  summarise(
    mean_val = mean(mean_N, na.rm = TRUE),
    se_val   = sd(mean_N, na.rm = TRUE) / sqrt(sum(!is.na(mean_N))),
    .groups = "drop"
  )

# --- Treatment main effect ---
treat_summary <- CN_5.10 |>
  group_by(Treatment) |>
  summarise(
    mean_val = mean(mean_N, na.rm = TRUE),
    se_val   = sd(mean_N, na.rm = TRUE) / sqrt(sum(!is.na(mean_N))),
    .groups = "drop"
  )

# --- Microsite main effect ---
micro_summary <- CN_5.10 |>
  group_by(Microsite) |>
  summarise(
    mean_val = mean(mean_N, na.rm = TRUE),
    se_val   = sd(mean_N, na.rm = TRUE) / sqrt(sum(!is.na(mean_N))),
    .groups = "drop"
  )

main_effects_summary <- bind_rows(
  site_summary   %>% mutate(effect = "Site"),
  treat_summary  %>% mutate(effect = "Treatment"),
  micro_summary  %>% mutate(effect = "Microsite")
)

# Step 6: Save output to a text file
sink(here::here("results", "anova_summary_pctN_5.10.txt"))
cat("=== Transformation Summary ===\n")
print(summary_table)
cat("\n=== ANOVA Summary for %N 5-10 cm depth ===\n")
print(anova_summary)
cat("\n=== Group Means and Standard Errors for trans_N and pctN ===\n")
print(group_summary)
print(group_summary1)
print(main_effects_summary)
sink()
```

```{r}
#| label: boxplot for 5-10 cm %N

# Microsite labels get screwed up in running this code multiple times - rerun from start 

# put the microsite, site, and treatment in order
CN_5.10 <- CN_5.10 |>
      mutate(
        Microsite = factor(Microsite, levels = c("Intercanopy", "Canopy"), 
                           labels = c("IC", "Canopy")),
        Treatment = factor(Treatment, levels = c("Exclusion","Ambient","Addition")),
        Site = factor(Site, levels = c("Desert Scrub","Desert Grassland",
                                       "Juniper Savanna", "Ponderosa Pine Meadow",
                                       "Mixed Conifer Meadow"))
        )

panel_labels <- data.frame(
  Site = unique(CN_5.10$Site),
  x = 0.6,  # Nudges left of the first x category
  y = 0.49,
  label = c("F.", "G.", "H.", "I.", "J.")
)

p_pctN_5.10 <- CN_5.10 |> 
  ggplot(aes(x = Microsite, y = mean_N, fill = Treatment)) + 
  geom_boxplot() +
  facet_wrap(~ Site, nrow = 1, labeller = labeller(Site = label_wrap_gen(width = 15))) +
  geom_text(
    data = panel_labels,
    inherit.aes = FALSE,
    aes(x = x, y = y, label = label),
    hjust = 0, vjust = 1,
    size = 5) +
  scale_y_continuous(limits = c(0, 0.5), breaks = c(0, 0.1, 0.2, 0.3, 0.4, 0.5), 
                     expand = c(0, 0)) +
  ylab("Total Nitrogen (%)") +
  xlab("Microsite") +
  theme_grey(base_size = 18) +
  scale_fill_manual(name = "", values = c("#de2d26", "#bdbdbd", "#08519c")) +
  theme_grey(base_size = 18) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "white", color = NA),
    axis.text.x = element_text(size = 14),
    axis.text = element_text(size = 15),
    panel.border = element_rect(color = "black", fill = NA),
    legend.position = "none") +
  guides(color = guide_legend(title = NULL))
p_pctN_5.10

ggsave(
  filename = here("figures", "pctN_5.10cm.png"),  
  plot = p_pctN_5.10,
  width = 12,       # width in inches
  height = 6,       # height in inches
  dpi = 300         # high-resolution PNG
)
```
