---
title: "InSituResp"
author: "Heather Throop"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

## Overview

This is a compilation, cleaning, and analysis of field-collected soil respiration data from Brittney Monus's MS research in the GRaMPS project.

Code was written by Heather Throop, streamlining and updating the original analyses that Brittney did for her MS.

Updated:

2025-07-09 Heather Throop\
2025-07-25 Heather Throop added soil C as s predictor of flux (needs some more work)

## SetUp

```{r}
#| label: load packages
library(tidyverse)
library(car)
library(MASS)
library(lme4)
library(here)
library(broom)
```

```{r}
#| label: prepare data

## read in L1 data and combine

# list of all CSV files in the folder
csv_files <- list.files(here("data", "L1_InSituRespiration"), 
                        pattern = "\\.csv$", full.names = TRUE)

# Read and combine all CSV files into one dataframe
InSituResp <- csv_files |>
  map_dfr(read_csv, show_col_types = FALSE) |>
  rename(RHcmbr = `RHcmbr%`) |> # get rid of the problematic "%"
  rename(Site = site) # make case consistent with other variables 

# Add a column for the season: pre-monsoon period (4-7 June, 18-21 June, and 2-5 July), peak-monsoon period (23-26 July and 14-17 August), and the post-monsoon period (11-14 September)
InSituResp <- InSituResp |>
  rename(SampleDate = date) |>
  mutate(Season = case_when(
           (SampleDate >= 180604 & SampleDate <= 180607) |
           (SampleDate >= 180618 & SampleDate <= 180621) |
           (SampleDate >= 180702 & SampleDate <= 180705) ~ "Pre",
           (SampleDate >= 180723 & SampleDate <= 180726) |
           (SampleDate >= 180814 & SampleDate <= 180817) ~ "Peak",
           (SampleDate >= 180911 & SampleDate <= 180914) ~ "Post",
           TRUE ~ NA_character_  # Set to NA if not in any defined range
         ))

# Add a column to separate out by sampling trip 
InSituResp <- InSituResp |>
  mutate(Trip = case_when(
           (SampleDate >= 180604 & SampleDate <= 180607) ~ "1",
           (SampleDate >= 180618 & SampleDate <= 180621) ~ "2",
           (SampleDate >= 180702 & SampleDate <= 180705) ~ "3",
           (SampleDate >= 180723 & SampleDate <= 180726) ~ "4",
           (SampleDate >= 180814 & SampleDate <= 180817) ~ "5",
           (SampleDate >= 180911 & SampleDate <= 180914) ~ "6",
           TRUE ~ NA_character_  # Set to NA if not in any defined range
         ))

## data clean-up

# Convert SampleDate to Date format
InSituResp <- InSituResp |>
  mutate(SampleDate = as.Date(as.character(SampleDate), format = "%y%m%d"))

# rename and order sites
InSituResp$Site = factor(InSituResp$Site, 
                              levels=c("BlackPoint","Antelope","BlueChute",
                                       "Arboretum","CampColton"), 
                              labels=c("Desert Scrub","Desert Grassland","Juniper Savanna",
                                       "Ponderosa Pine Meadow","Mixed Conifer Meadow")) 
# rename and order treatments
InSituResp$Treatment = factor(InSituResp$Treatment, 
                                   levels=c("Exl","Ctrl","Add"), 
                                   labels=c("Exclusion","Ambient","Addition")) 

# rename canopy location
InSituResp$Microsite = factor(InSituResp$Microsite, 
                                   levels=c("IC", "Can"), 
                                   labels=c("Intercanopy", "Canopy")) 

# put the seasons in order
InSituResp <- InSituResp |>
      mutate(Season = factor(Season, levels = c("Pre", "Peak", "Post")))
```

## Daily Means

There are two collars in each plot - one per microsite. Multiple measurements were taken at a time for each collar using the protocol. Here we reduce the data to the mean for each collar (still two collars per plot).

```{r}
#| label: daily collar means

# calculate collar-level means for replicate respiration measurements per collar (~ 4 replicate Licor cycles per collar) for each date x plot x site combination x treatment x microsite combination
InSituResp_daily_collar_means <- InSituResp |>
  group_by(SampleDate, Site, Plot, Treatment, Microsite, Season, Trip) |>
  summarise(
    N_reps    = sum(!is.na(EFFLUX)),
    Efflux   = mean(EFFLUX, na.rm = TRUE),
    RHcmbr   = mean(RHcmbr, na.rm = TRUE),
    Tsoil_C   = mean(Tsoil_C, na.rm = TRUE), 
    Tsch_C   = mean(Tsch_C, na.rm = TRUE),
    .groups = "drop"
  )
```

```{r}
#| label: graph plot x microsite means for each trip

p_trt_x_microsite <- InSituResp_daily_collar_means |> 
  ggplot(aes(x = SampleDate, y = Efflux, color = Treatment)) + 
  geom_point(position = position_dodge(width = 0.5), size = 2) +
  facet_grid(rows = vars(Site), cols = vars(Microsite), labeller = labeller(
    Site = label_wrap_gen(width = 15),
    Microsite = label_wrap_gen(width = 15))) +
  scale_y_continuous(limits = c(0, 7.5), breaks = c(0, 2, 4, 6)) +
  scale_x_date(date_breaks = "2 weeks", date_labels = "%b %d") +
  ylab(expression(paste(Soil~CO[2]~Efflux~(mu*mol*CO[2]*m^{-2}*s^{-1})))) +
  xlab("Sample Date") +
  theme_grey(base_size = 18) +
  scale_color_manual(name = "Treatment", values = c("#de2d26", "#bdbdbd", "#08519c")) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 90, hjust = 1, size = 10),
    axis.text = element_text(size = 15),
    panel.border = element_rect(color = "black", fill = NA),
    legend.position = "top"
  ) +
  guides(color = guide_legend(title = "Treatment"))
p_trt_x_microsite

```

## Seasonal Means

Here the data are plotted by three seasons (pre-, peak, and post-monsoon), merging the multiple sampling trips that occurred in the pre-monsoon and peak-monsoon seasons.

```{r}
#| label: boxplot by season, separate microsites

# separating out by microsite
p_trt_x_microsite <- InSituResp_daily_collar_means |> 
  ggplot(aes(x = factor(Season), y = Efflux, fill = Treatment, Microsite)) + 
  geom_boxplot(position = position_dodge(width = 0.8), outlier.shape = NA) +
 facet_grid(cols = vars(Site), rows = vars(Microsite), labeller = labeller(
    site = label_wrap_gen(width = 15),
    Microsite = label_wrap_gen(width = 15))) +
  scale_y_continuous(limits = c(0, 7.5), breaks = c(0, 2, 4, 6)) +
  ylab(expression(paste(Soil~CO[2]~Efflux~(mu*mol*CO[2]*m^{-2}*s^{-1})))) +
  xlab("Season") +
  theme_grey(base_size = 18) +
#  scale_fill_manual(name = "Treatment", values = c("#de2d26", "#bdbdbd", "#08519c")) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 14),
    axis.text = element_text(size = 15),
    panel.border = element_rect(color = "black", fill = NA),
    legend.position = "top"
  ) +
  guides(fill = guide_legend(title = "Treatment"))
p_trt_x_microsite

```

Merged the microsites together for a simpler figure.

```{r}
#| label: boxplot: efflux by seasoen (microsites combimned)

# microsites merged

# Create annotation data frame for facet labels
panel_labels <- data.frame(
  Site = unique(InSituResp_daily_collar_means$Site),
  x = 0.6,  # Nudges left of the first x category
  y = 7.3,
  label = c("C.", "B.", "D.", "A.", "E.")
)

# Plot
p_trt_x_season <- InSituResp_daily_collar_means |> 
  ggplot(aes(x = Season, y = Efflux, fill = Treatment)) + 
  geom_boxplot(position = position_dodge(width = 0.8), outlier.shape = NA) +
  facet_wrap(~Site, ncol = 5, labeller = labeller(Site = label_wrap_gen(width = 20))) +
 geom_text(
  data = panel_labels,
  inherit.aes = FALSE,
  aes(x = x, y = y, label = label),
  hjust = 0, vjust = 1,
  size = 5) +
  scale_y_continuous(limits = c(0, 7.5), breaks = c(0, 2, 4, 6), expand = c(0, 0)) +
  ylab(expression(paste(Soil~CO[2]~Efflux~(mu*mol*~CO[2]*~m^{-2}*~s^{-1})))) +
  xlab("Monsoon Season") +
  scale_fill_manual(name = "", values = c("#de2d26", "#bdbdbd", "#08519c")) +
  theme_grey(base_size = 18) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "white", color = NA),
    axis.ticks.x = element_blank(),
    axis.text.x = element_text(size = 14),
    axis.text = element_text(size = 15),
    panel.border = element_rect(color = "black", fill = NA),
    legend.position = c(0.02, 0.88),
    legend.justification = c("left", "top")
   )

p_trt_x_season

ggsave(
  filename = here("figures", "ExSituFlux_by_season.png"),  
  plot = p_trt_x_season,
  width = 12,       # width in inches
  height = 6,       # height in inches
  dpi = 300         # high-resolution PNG
)
```

## Drivers of InSitu Respiration

Explore potential drivers of in situ respiration (moisture, ANPP, and soil CN)

This combines the dataframe InSituResp_daily_collar_means with dailymoisturemeans_byplot\
Uses model selection to come up with predictive relationship for InSituResp

```{r}
#| label: combine insitu resp and moisture data

# combine the daily mean values for VWC (by plot) with mean values for efflux (by plot and microsite)

# read in the VWC data
dailymoisturemeans_byplot <- read_csv(here("data","calculated_outputs","dailymoisturemeans_by_plot.csv"))

# Ensure columns are properly coded to combine df
dailymoisturemeans_byplot <- dailymoisturemeans_byplot |>
  mutate(across(c(Site, Plot, Treatment), as.factor)) 

InSituResp_daily_collar_means$Plot <- as.factor(InSituResp_daily_collar_means$Plot)

# note that there are two efflux values from each plot x day combo (the two microsites) for efflux but just one VWC value
# some plots did not have soil moisture probes, so these plots have NA for VWC
InSituResp_combined <- InSituResp_daily_collar_means |>
  left_join(
    dailymoisturemeans_byplot |>
      dplyr::select(SampleDate, Site, Plot, Treatment, 
             SWP = mean_SWP, VWC = mean_VWC),
    by = c("SampleDate", "Site", "Plot", "Treatment")
  )
```

```{r}
#| label: regression: plot-level VWC vs efflux
# regression with plot-level VWC, with mean efflux values the mean of the two microsites for each plot
# Use mean site x treatment x date values for VWC to fill in missing VWC values (where there is no soil moisture probe)

InSituResp_combined_plotmeans <- InSituResp_combined |>
  group_by(SampleDate, Site, Plot, Treatment, Season, Trip) |>
  summarise(
    N_reps    = sum(!is.na(Efflux)),
    Efflux   = mean(Efflux, na.rm = TRUE),
    RHcmbr   = mean(RHcmbr, na.rm = TRUE),
    Tsoil_C   = mean(Tsoil_C, na.rm = TRUE), 
    Tsch_C   = mean(Tsch_C, na.rm = TRUE),
    SWP = mean(SWP, na.rm = TRUE),
    VWC = mean(VWC, na.rm = TRUE),
    .groups = "drop"
  )

# replace NaN with mean VWC for each site x treatment x sampledate combination
InSituResp_combined_plotmeans <- InSituResp_combined_plotmeans |>
  group_by(Site, Treatment, SampleDate) |>
  mutate(VWC = ifelse(is.na(VWC), mean(VWC, na.rm = TRUE), VWC)) |>
  ungroup()

# regression model for each site (note that this is oddly printed with two lines per site)
flux_v_VWC_regression_results <- InSituResp_combined_plotmeans |>
  group_by(Site) |>
  nest() |>
  mutate(
    model = map(data, ~ lm(Efflux ~ VWC, data = .x)),
    model_summary = map(model, tidy),
    model_stats = map(model, glance)
  ) |>
  unnest(model_summary, names_sep = "_") |>
  pivot_wider(
    names_from = model_summary_term,
    values_from = model_summary_estimate
  ) |>
  unnest(model_stats, names_sep = "_") |>
  dplyr::select(
    Site,
    intercept = `(Intercept)`,
    slope = VWC,
    r_squared = model_stats_r.squared,
    p_value = model_stats_p.value
  )

flux_v_VWC_regression_results
```

```{r}
#|label: graph: efflux vs VWC

# Create annotation data frame for facet labels
panel_labels <- data.frame(
  Site = unique(InSituResp_combined_plotmeans$Site),
  x = 0.01,  # Nudges left of the first x category
  y = 8.45,
  label = c("H.", "F.", "I.", "F.", "J.")
)

# make the plot
resp_v_VWC_p1 <- InSituResp_combined_plotmeans |>
  ggplot(aes(x= VWC, y=Efflux, color = Treatment)) + 
  geom_point(position = "identity", size = 2) +
  #facet_grid(.~Site) +
  facet_wrap(~Site, ncol = 5, labeller = labeller(Site = label_wrap_gen(width = 20))) +
  geom_text(
  data = panel_labels,
  inherit.aes = FALSE,
  aes(x = x, y = y, label = label),
  hjust = 0, vjust = 1,
  size = 5) +
  stat_smooth(aes(group = 1), color="black", method = "lm", se = FALSE, formula = y ~ x) +
  ylab(expression(paste(Soil~CO[2]~Efflux~(mu*mol*~CO[2]*~m^{-2}*~s^{-1})))) +
  xlab(expression(paste(Volumetric~Water~Content~(m^{3}~water~m^{-3}~soil))))+
  theme_grey(base_size = 18) +
  scale_shape_manual(values=c(15,16,17,0,4), name = "Site") +
  scale_color_manual(values=c("#de2d26","#bdbdbd","#08519c"  ),labels = c ("Exclusion","Ambient","Addition"))+
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.ticks.x=element_blank(),
    legend.position = "none",
    axis.text.x = element_text(size = 14),
    axis.text = element_text(size = 15),
    panel.background = element_blank(),
    panel.border = element_rect(color = "black", fill = NA)) 
resp_v_VWC_p1

ggsave(here("figures", "ExSituFlux_vs_VWC.png"), plot = resp_v_VWC_p1, width = 10, height = 8)
```

```{r}
#| label: combine insitu resp and CN data

## read in L2 CN data - there is one value for each site x treatment x plot x microsite x depth combination
CN_L2 = read.csv(here("data","calculated_outputs","CN_L2.csv"))

# calculate CN core means (0-10 cm) for across the two soils depths for each plot x site x treatment x microsite combination
CN_core_means <- CN_L2 |>
  group_by(Site, Plot, Treatment, Microsite) |>
  summarise(
    N    = n(),
    pctC_0.10   = mean(mean_C, na.rm = TRUE),
    pctN_0.10   = mean(mean_N, na.rm = TRUE),
    .groups = "drop"
  ) |>
  filter(N == 2) # remove values when we have only 1 of the 2 depths

# Ensure columns are properly coded as factors to allow combining of dfs
CN_core_means <- CN_core_means |>
  mutate(across(c(Site, Plot, Treatment, Microsite), as.factor)) 

# combine CN data (means of 0-5 and 5-10 cm) and resp data
InSituResp_combined <- InSituResp_combined |>
  left_join(
    CN_core_means |>
      dplyr::select(Site, Plot, Treatment, Microsite, pctC_0.10, pctN_0.10),
    by = c("Site", "Plot", "Treatment", "Microsite")
  )

## combine CN 0-5 and 5-10 cm and resp data
# prepare depth-specific data for the join
CN_0.5 <- CN_L2 |>
  filter(Depth == "0.5") |>
  rename(C_0.5 = mean_C) |>
  rename(N_0.5 = mean_N) |>
  mutate(across(c(Site, Plot, Treatment, Microsite), as.factor)) 
CN_5.10 <- CN_L2 |>
  filter(Depth == "5.1") |>
  rename(C_5.10 = mean_C) |>
  rename(N_5.10 = mean_N) |>
  mutate(across(c(Site, Plot, Treatment, Microsite), as.factor)) 

# do the join, adding in plot and microsite-specific values for C and N at 0-5 and 5-10 cm depth
InSituResp_combined <- InSituResp_combined |>
  left_join(
    CN_0.5 |>
      dplyr::select(Site, Plot, Treatment, Microsite, C_0.5, N_0.5),
    by = c("Site", "Plot", "Treatment", "Microsite")
  )
InSituResp_combined <- InSituResp_combined |>
  left_join(
    CN_5.10 |>
      dplyr::select(Site, Plot, Treatment, Microsite, C_5.10, N_5.10),
    by = c("Site", "Plot", "Treatment", "Microsite")
  )

### DO I NEED TO DO SOME GAP FILLING FOR MISSING CN VALUES?


```

```{r}
#| label: combine insitu resp and ANPP data

## read in L2 ANPP and root biomass data - there is one value for each site x treatment x plot combination for ANPP. Roots mass is separated out by microsite type. 
ANPP_L2 = read.csv(here("data", "calculated_outputs", "ANPPandRootBiomass_L2.csv"))

# convert ANPP to long form data to pull out root biomass by microsite
ANPP_long <- ANPP_L2 |>
  pivot_longer(
    cols = c(Root_mass_IC_gm2, Root_mass_Can_gm2),
    names_to = "Microsite",
    values_to = "Root_mass_gm2"
  ) |>
  mutate(
    Microsite = case_when(
      Microsite == "Root_mass_IC_gm2" ~ "Intercanopy",
      Microsite == "Root_mass_Can_gm2" ~ "Canopy",
      TRUE ~ Microsite
    )
  ) |>
  dplyr::select(SiteOrder, TreatmentOrder, Plot, Microsite, Root_mass_gm2, plot_mean_ANPP)

# Ensure columns are properly coded as factors to allow combining of dfs
ANPP_long <- ANPP_long |>
  mutate(across(c(SiteOrder, Plot, TreatmentOrder, Microsite), as.factor)) |>
  rename(Site = SiteOrder, Treatment = TreatmentOrder)

# combine ANPP and root data with resp data
InSituResp_combined <- InSituResp_combined |>
  left_join(
    ANPP_long |>
      dplyr::select(Site, Plot, Treatment, Microsite, plot_mean_ANPP, Root_mass_gm2),
    by = c("Site", "Plot", "Treatment", "Microsite")
  )
```

```{r}
#| label: data gap filling

# There are missing values for many of the predictor values (e.g., missing CN values, soil moisture probes not present in all plots). This will cause problems due to unequal sample sizes. Gap fill with the mean Site x Treatment x Microsite value.


InSituResp_combined <- InSituResp_combined |>
  group_by(Site, Treatment, Microsite) |>
  mutate(
    VWC = if_else(is.na(VWC), mean(VWC, na.rm = TRUE), VWC),
    C_0.5 = if_else(is.na(C_0.5), mean(C_0.5, na.rm = TRUE), C_0.5),
    plot_mean_ANPP = if_else(is.na(plot_mean_ANPP), mean(plot_mean_ANPP, na.rm = TRUE), plot_mean_ANPP),
    Root_mass_gm2 = if_else(is.na(Root_mass_gm2), mean(Root_mass_gm2, na.rm = TRUE), Root_mass_gm2)
  ) |>
  ungroup()
```

```{r}
#| label: model: predictive relationship for InSituResp

# Use linear mixed effects models to predict Efflux based on VWC and Tsoil_C, with a different model for each site
library(lme4)

# Make sure Site and Plot are factors
InSituResp_combined <- InSituResp_combined |>
  mutate(
    Site = as.factor(Site),
    Plot = as.factor(Plot)
  )

# Fit different mixed effects model and compare model fits with AIC
library(lme4)
library(lmerTest)  # for stepwise methods with p-values (optional)
library(performance)


# Full model with all fixed effects
full_model <- lmer(
  Efflux ~ VWC + Treatment + Site + Tsoil_C + C_0.5 + 
    Microsite + plot_mean_ANPP + Root_mass_gm2 + (1 | Plot),
  data = InSituResp_combined,
  REML = FALSE
)


# Uses Satterthwaite's method to calculate p-values
stepwise_model <- step(full_model)
summary(stepwise_model)

final_model <- get_model(stepwise_model)
summary(final_model)

sink(here("results","insitu_resp_model_summary.txt"))
summary(final_model)
sink()

```
