---
title: "Microbial Biomass C"
author: "Heather Throop"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

### Overview

This is a compilation, cleaning, and analysis of microbial biomass C data from Brittney Monus's MS research in the GRaMPS project.

MBC = extractable.MBC/0.45 per Jenkinson et al. (2004). Soil Biology and Biochemistry 36(1): 5-7. Some people think this number is somewhat arbitrary and prefer report extractable MBC.\
All concentrations are in mg C per kg dry soil

Code was written by Heather Throop, streamlining and updating the original analyses that Brittney did for her MS.

Written:\
2025-07-27 Heather Throop\
Revised:\
2025-10-07 Heather Throop - minor figure revisions

## SetUp

```{r}
#| label: load packages
library(tidyverse)
library(car)       # For Levene's test
library(emmeans)   # For post-hoc comparisons
library(here)
library(bestNormalize)
library(cowplot)
library(grid)
```

## Ingest and Clean L1 Data

This chunk can be skipped - start with ingesting L2 data (below)

```{r}
#| label: prepare data

# list of all CSV files in the folder
csv_files <- list.files(here("data","L1_MicrobialBiomass"), 
                        pattern = "\\.csv$", full.names = TRUE)

# Read and combine all CSV files into one dataframe
MicrobBio <- csv_files |>
  map_dfr(read_csv, show_col_types = FALSE) 

# rename and order sites, treatments, and microsites
MicrobBio <- MicrobBio |>
      mutate(Site = factor(Site, levels = c("BP","ANT","BC","ARB","CC"), 
                              labels = c("Desert Scrub","Desert Grassland",
                                       "Juniper Savanna",
                                       "Ponderosa Pine Meadow","Mixed Conifer Meadow")),
             Treatment = factor(Treatment, levels = c("Exl", "Add", "Ctrl"),
                                labels = c("Exclusion","Ambient","Addition")),
             Microsite = factor(Microsite, 
                                   levels=c("IC", "Can"), 
                                   labels=c("Intercanopy", "Canopy")),
             Plot = factor(Plot),
             Depth = factor(Depth))

# Clean data to remove 0 values for extractable.MBC and MBC
MicrobBio <- MicrobBio |>
  filter(extractable.MBC > 0)

### reduce to depth x plot means
# calculate mean and SE for each Site x Treatment x Microsite x Depth x Plot combination. In most cases there was only one analytical replicate per plot, but this accounts for cases where there were analytical reps.
MicrobBio_clean <- MicrobBio |>
  group_by(Site, Treatment, Microsite, Depth, Plot) |>
  summarise(
    N = n(),
    extractable.MBC = mean(extractable.MBC, na.rm = TRUE),
    MBC = mean(MBC, na.rm = TRUE),
    .groups = 'drop'
  )

# save clean L2 data ready for analysis
write.csv(MicrobBio_clean, 
          here("data","calculated_outputs","MicrobialBiomass_L2.csv"), row.names = FALSE)

```

## MBC patterns by Depth

Start here to pull in cleaned L2 data

```{r}
#| label: load L2 data

## read in L2 data - one value for each site x treatment x plot x microsite x depth combination
MBC_df = read.csv(here("data","calculated_outputs","MicrobialBiomass_L2.csv"))

# Ensure factors are properly coded
MBC_df <- MBC_df |>
  mutate(across(c(Site, Treatment, Plot, Microsite, Depth), as.factor))
```

### 0-5 cm depth

#### microsite x site x treatment

##### ANOVA

```{r}
#| label: Check ANOVA assumptions - MBC 0-5 cm 

# Filter and prepare data
MBC_0.5_df <- MBC_df |>
  filter(Depth == "upper") |> 
  mutate(Group = interaction(Site, Treatment, Microsite))

# visualize outliers
ggplot(MBC_0.5_df, aes(x = Group, y = MBC)) +
  geom_boxplot(outlier.color = "red") +
  theme_bw()
geom_jitter(width = 0.2, alpha = 0.6)

# Remove outliers based on 1.5*IQR rule within each group
MBC_0.5_outliers_rmvd <- MBC_0.5_df |>
  group_by(Group) |>
  filter(between(MBC,
                 quantile(MBC, 0.25) - 1.5 * IQR(MBC),
                 quantile(MBC, 0.75) + 1.5 * IQR(MBC))) |>
  ungroup()

# Check initial normality
hist(MBC_0.5_outliers_rmvd$MBC, main = "Histogram of mean_C (0–5 cm)", xlab = "mean_C")
qqnorm(MBC_0.5_outliers_rmvd$MBC); qqline(MBC_0.5_outliers_rmvd$MBC)

shapiro_raw <- shapiro.test(MBC_0.5_outliers_rmvd$MBC)
cat("Shapiro-Wilk p-value (raw):", shapiro_raw$p.value, "\n")

# Apply bestNormalize if needed
if (shapiro_raw$p.value < 0.05) {
  bn_obj <- bestNormalize(MBC_0.5_outliers_rmvd$MBC, allow_orderNorm = FALSE)
  MBC_0.5_outliers_rmvd <- MBC_0.5_outliers_rmvd |>
    mutate(MBC_trans = predict(bn_obj, newdata = MBC))
  used_var <- "MBC_trans"
  transform_used <- class(bn_obj$chosen_transform)[1]
  
# Check transformed normality
  hist(MBC_0.5_outliers_rmvd$MBC_trans, main = "Histogram of Transformed C", xlab = "MBC_trans")
  qqnorm(MBC_0.5_outliers_rmvd$MBC_trans); qqline(MBC_0.5_outliers_rmvd$MBC_trans)
  
  shapiro_trans <- shapiro.test(MBC_0.5_outliers_rmvd$MBC_trans)
  cat("Shapiro-Wilk p-value (transformed):", shapiro_trans$p.value, "\n")
} else {
  used_var <- "MBC"
  transform_used <- "None"
}

# Levene's Test
levene_result <- leveneTest(reformulate("Group", used_var), data = MBC_0.5_outliers_rmvd)

# Summary Table
summary_table <- tibble(
  variable = used_var,
  transformed = transform_used,
  shapiro_p_raw = shapiro_raw$p.value,
  shapiro_p_trans = if (exists("shapiro_trans")) shapiro_trans$p.value else NA,
  levene_p = levene_result$`Pr(>F)`[1]
)
print(summary_table)
```

```{r}
#| label: three-way ANOVA

# Three-way ANOVA model
anova_model <- aov(reformulate(c("Site", "Treatment", "Microsite",
                                  "Site:Treatment",
                                  "Site:Microsite",
                                  "Treatment:Microsite",
                                  "Site:Treatment:Microsite"),
                                used_var),
                   data = MBC_0.5_outliers_rmvd)

# Summarize ANOVA
anova_summary <- summary(anova_model)
print(anova_summary)

# Pairwise comparisons (Tukey's HSD)
tukey_results <- TukeyHSD(anova_model)

# Step 6: Save output to a text file
sink(here::here("results", "anova_summary_MBC_0.5_df.txt"))
cat("=== Transformation Summary ===\n")
print(summary_table)

cat("\n=== ANOVA Summary ===\n")
print(anova_summary)

cat("\n=== Pairwise Comparisons (Tukey's HSD) ===\n")
for (term in names(tukey_results)) {
  cat("\n---", term, "---\n")
  print(tukey_results[[term]])
}
sink()

```

##### Boxplots

```{r}
#| label: boxplot for 0-5 cm MBC

# put the microsite, site, and treatment in order
MBC_0.5_outliers_rmvd <- MBC_0.5_outliers_rmvd |>
      mutate(
        Microsite = factor(Microsite, levels = c("Intercanopy","Canopy")),
        Treatment = factor(Treatment, levels = c("Exclusion","Ambient","Addition")),
        Site = factor(Site, levels = c("Desert Scrub","Desert Grassland",
                                       "Juniper Savanna", "Ponderosa Pine Meadow",
                                       "Mixed Conifer Meadow"))
        )

panel_labels <- data.frame(
  Site = unique(MBC_0.5_df_outliers_rmvd$Site),
  x = 0.6,  # Nudges left of the first x category
  y = 1150,
  label = c("A.", "B.", "C.", "D.", "E.")
)

p_MBC_0.5 <- MBC_0.5_outliers_rmvd |> 
  ggplot(aes(x = Microsite, y = MBC, fill = Treatment)) + 
  geom_boxplot() +
  facet_wrap(~ Site, nrow = 1, labeller = labeller(Site = label_wrap_gen(width = 15))) +
    geom_text(
    data = panel_labels,
    inherit.aes = FALSE,
    aes(x = x, y = y, label = label),
    hjust = 0, vjust = 1,
    size = 5) +
  labs(
    x = "Microsite",
    y = expression(Microbial~Biomass~(mg~DOC~kg^{-1}~dry~soil))) +
  theme_grey(base_size = 18) +
  scale_fill_manual(name = "", values = c("#de2d26", "#bdbdbd", "#08519c")) +
  theme_grey(base_size = 18) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "white", color = NA),
    axis.text.x = element_text(size = 14),
    axis.text = element_text(size = 15),
    panel.border = element_rect(color = "black", fill = NA),
    legend.position = c(0.002, 0.88),
    legend.justification = c("left", "top")
  ) +
   scale_y_continuous(limits = c(0, 1200), expand = c(0, 0)) +
  guides(color = guide_legend(title = NULL))
p_MBC_0.5

ggsave(
  filename = here("figures", "MBC_0.5cm.png"),  
  plot = p_MBC_0.5,
  width = 12,       # width in inches
  height = 6,       # height in inches
  dpi = 300         # high-resolution PNG
)
```

#### site x microsite

```{r}
#| label: boxplot for 0-5 cm MBC - treatments combined

p_MBC_0.5_treatments_combined <- MBC_0.5_df |> 
  ggplot(aes(x = Site, y = MBC, fill = Microsite)) + 
  geom_boxplot() +
  labs(
    x = "Site",
    y = expression(Microbial~Biomass~(mg~DOC~kg^{-1}~dry~soil))) +
  theme_grey(base_size = 18) +
  scale_fill_manual(name = "", values = c("white", "grey")) +
  theme_grey(base_size = 18) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "white", color = NA),
    axis.text.x = element_text(size = 14),
    axis.text = element_text(size = 15),
    panel.border = element_rect(color = "black", fill = NA),
    legend.position = c(0.002, 0.88),
    legend.justification = c("left", "top")
  ) +
   scale_y_continuous(limits = c(0, 750), expand = c(0, 0))# +
   guides(color = guide_legend(title = NULL))
p_MBC_0.5_treatments_combined

ggsave(
  filename = here("figures", "MBC_0.5cm_treatments_combined.png"),  
  plot = p_MBC_0.5_treatments_combined,
  width = 12,       # width in inches
  height = 6,       # height in inches
  dpi = 300         # high-resolution PNG
)
```

### 5-10 cm depth

#### site x treatment x microsite

```{r}
#| label: Check ANOVA assumptions - MBC 5-10 cm 

# Filter and prepare data
MBC_5.10_df <- MBC_df |>
  filter(Depth == "lower") |>
  mutate(Group = interaction(Site, Treatment, Microsite))

# visualize outliers
ggplot(MBC_5.10_df, aes(x = Group, y = MBC)) +
  geom_boxplot(outlier.color = "red") +
  theme_bw()
geom_jitter(width = 0.2, alpha = 0.6)

# Remove outliers based on 1.5*IQR rule within each group
MBC_5.10_outliers_rmvd <- MBC_5.10_df |>
  group_by(Group) |>
  filter(between(MBC,
                 quantile(MBC, 0.25) - 1.5 * IQR(MBC),
                 quantile(MBC, 0.75) + 1.5 * IQR(MBC))) |>
  ungroup()

# Check initial normality
hist(MBC_5.10_outliers_rmvd$MBC, main = "Histogram of mean_C (0–5 cm)", xlab = "mean_C")
qqnorm(MBC_5.10_outliers_rmvd$MBC); qqline(MBC_5.10_outliers_rmvd$MBC)

shapiro_raw <- shapiro.test(MBC_5.10_outliers_rmvd$MBC)
cat("Shapiro-Wilk p-value (raw):", shapiro_raw$p.value, "\n")

# Apply bestNormalize if needed
if (shapiro_raw$p.value < 0.05) {
  bn_obj <- bestNormalize(MBC_5.10_outliers_rmvd$MBC, allow_orderNorm = FALSE)
  MBC_5.10_outliers_rmvd <- MBC_5.10_outliers_rmvd |>
    mutate(MBC_trans = predict(bn_obj, newdata = MBC))
  used_var <- "MBC_trans"
  transform_used <- class(bn_obj$chosen_transform)[1]
  
  # Check transformed normality
  hist(MBC_5.10_outliers_rmvd$MBC_trans, main = "Histogram of Transformed C", xlab = "MBC_trans")
  qqnorm(MBC_5.10_outliers_rmvd$MBC_trans); qqline(MBC_5.10_outliers_rmvd$MBC_trans)
  
  shapiro_trans <- shapiro.test(MBC_5.10_outliers_rmvd$MBC_trans)
  cat("Shapiro-Wilk p-value (transformed):", shapiro_trans$p.value, "\n")
} else {
  used_var <- "MBC"
  transform_used <- "None"
}

# Levene's Test
levene_result <- leveneTest(reformulate("Group", used_var), data = MBC_5.10_outliers_rmvd)

# Summary Table
summary_table2 <- tibble(
  variable = used_var,
  transformed = transform_used,
  shapiro_p_raw = shapiro_raw$p.value,
  shapiro_p_trans = if (exists("shapiro_trans")) shapiro_trans$p.value else NA,
  levene_p = levene_result$`Pr(>F)`[1]
)
print(summary_table2)
```

```{r}
#| label: three-way ANOVA

# Three-way ANOVA model
anova_model2 <- aov(reformulate(c("Site", "Treatment", "Microsite",
                                  "Site:Treatment",
                                  "Site:Microsite",
                                  "Treatment:Microsite",
                                  "Site:Treatment:Microsite"),
                                used_var),
                   data = MBC_5.10_outliers_rmvd)
anova_summary2 <- summary(anova_model2)
print(anova_summary2)

# Pairwise comparisons (Tukey's HSD)
tukey_results <- TukeyHSD(anova_model2)

# Step 6: Save output to a text file
sink(here::here("results", "anova_summary_MBC_5.10_df.txt"))
cat("=== Transformation Summary ===\n")
print(summary_table2)

cat("\n=== ANOVA Summary ===\n")
print(anova_summary2)

cat("\n=== Pairwise Comparisons (Tukey's HSD) ===\n")
for (term in names(tukey_results)) {
  cat("\n---", term, "---\n")
  print(tukey_results[[term]])
}
sink()

```

```{r}
#| label: boxplot for 5-10 cm MBC

# put the microsite, site, and treatment in order
MBC_5.10_outliers_rmvd <- MBC_5.10_outliers_rmvd |>
      mutate(
        Microsite = factor(Microsite, levels = c("Intercanopy","Canopy")),
        Treatment = factor(Treatment, levels = c("Exclusion","Ambient","Addition")),
        Site = factor(Site, levels = c("Desert Scrub","Desert Grassland",
                                       "Juniper Savanna", "Ponderosa Pine Meadow",
                                       "Mixed Conifer Meadow"))
        )

panel_labels <- data.frame(
  Site = unique(MBC_5.10_outliers_rmvd$Site),
  x = 0.6,  # Nudges left of the first x category
  y = 725,
  label = c("F.", "G.", "H.", "I.", "J.")
)

p_MBC_5.10 <- MBC_5.10_outliers_rmvd |> 
  ggplot(aes(x = Microsite, y = MBC, fill = Treatment)) + 
  geom_boxplot() +
  facet_wrap(~ Site, nrow = 1, labeller = labeller(Site = label_wrap_gen(width = 15))) +
    geom_text(
    data = panel_labels,
    inherit.aes = FALSE,
    aes(x = x, y = y, label = label),
    hjust = 0, vjust = 1,
    size = 5) +
 labs(
    x = "Site",
    y = expression(Microbial~Biomass~(mg~DOC~kg^{-1}~dry~soil))) +
  theme_grey(base_size = 18) +
  scale_fill_manual(name = "", values = c("#de2d26", "#bdbdbd", "#08519c")) +
  theme_grey(base_size = 18) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "white", color = NA),
    axis.text.x = element_text(size = 14),
    axis.text = element_text(size = 15),
    panel.border = element_rect(color = "black", fill = NA),
    legend.position = c(0.002, 0.88),
    legend.justification = c("left", "top")
  ) +
   scale_y_continuous(limits = c(0, 750), expand = c(0, 0)) +
  guides(color = guide_legend(title = NULL))
p_MBC_5.10

ggsave(
  filename = here("figures", "MBC_5.10cm.png"),  
  plot = p_MBC_5.10,
  width = 12,       # width in inches
  height = 6,       # height in inches
  dpi = 300         # high-resolution PNG
)
```

#### site x microsite

```{r}
#| label: boxplot for 5-10 cm MBC - treatments combined

p_MBC_5.10_treatments_combined <- MBC_5.10_df |> 
  ggplot(aes(x = Site, y = MBC, fill = Microsite)) + 
  geom_boxplot() +
  labs(
    x = "Site",
    y = expression(Microbial~Biomass~(mg~DOC~kg^{-1}~dry~soil))) +
  theme_grey(base_size = 18) +
  scale_fill_manual(name = "", values = c("white", "grey")) +
  theme_grey(base_size = 18) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "white", color = NA),
    axis.text.x = element_text(size = 14),
    axis.text = element_text(size = 15),
    panel.border = element_rect(color = "black", fill = NA),
    legend.position = c(0.002, 0.88),
    legend.justification = c("left", "top")
  ) +
   scale_y_continuous(limits = c(0, 750), expand = c(0, 0)) +
  guides(color = guide_legend(title = NULL))
p_MBC_5.10_treatments_combined

ggsave(
  filename = here("figures", "MBC_5.10cm_treatments_combined.png"),  
  plot = p_MBC_5.10_treatments_combined,
  width = 12,       # width in inches
  height = 6,       # height in inches
  dpi = 300         # high-resolution PNG
)
```

### Combined depths

```{r}
#| label: combined depth boxplot - treatments included

# Customize x-axis labels to add second line when needed
site_labels <- c(
  "Desert Scrub" = "Desert Scrub",
  "Desert Grassland" = "Desert Grassland",
  "Juniper Savanna" = "Juniper Savanna",
  "Ponderosa Pine Meadow" = "Ponderosa Pine\nMeadow",
  "Mixed Conifer Meadow" = "Mixed Conifer\nMeadow"  
)

# Remove x-axis and y-axis titles from top plot; keep legend and add label
p_top1 <- p_MBC_0.5 +
  theme(
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_line(),
    legend.position = c(0.002, 0.88),
    legend.justification = c("left", "top")
  ) +
  annotate("text", x = -Inf, y = Inf, label = "", 
           hjust = -0.1, vjust = 1.5, size = 6, fontface = "plain") +
  scale_x_discrete(labels = site_labels)

# Remove both axis titles and legend from bottom plot; add label
p_bottom1 <- p_MBC_5.10 +
  theme(
    axis.title.y = element_blank(),
    axis.title.x = element_text(size = 16),
    legend.position = "none",
    strip.text.x = element_blank(),       # removes facet labels
    strip.background = element_blank()    # removes the gray background strip
  ) +
  annotate("text", x = -Inf, y = Inf, label = "", 
           hjust = -0.1, vjust = 1.5, size = 6, fontface = "plain") +
  scale_x_discrete(labels = site_labels)

# Combine vertically
combined_plot1 <- plot_grid(
  p_top1, p_bottom1,
  ncol = 1,
  align = "v",
  axis = "lr",
  rel_heights = c(1, 1)
)

# Draw final plot with shared y-axis label
# Add left margin space using a blank spacer
final_plot1 <- ggdraw() +
  draw_label(
    expression("Microbial Biomass Carbon (mg DOC kg"^-1*" soil)"),
    x = 0.02, y = 0.5, angle = 90, size = 18,
    fontface = "plain"
  ) +
  draw_plot(combined_plot1, x = 0.05, y = 0, width = 0.95, height = 1)

# Display the final plot
final_plot1

ggsave(
  filename = here("figures", "MBC_by_trt_combined_depth_panels.png"),  
  plot = final_plot1,
  width = 12,       # width in inches
  height = 6,       # height in inches
  dpi = 300         # high-resolution PNG
)
```

```{r}
#| label: combined depth boxplot - no treatments included

# Customize x-axis labels to add second line when needed
site_labels <- c(
  "Desert Scrub" = "Desert Scrub",
  "Desert Grassland" = "Desert Grassland",
  "Juniper Savanna" = "Juniper Savanna",
  "Ponderosa Pine Meadow" = "Ponderosa Pine\nMeadow",
  "Mixed Conifer Meadow" = "Mixed Conifer\nMeadow"  
)

# Remove x-axis and y-axis titles from top plot; keep legend and add label
p_top <- p_MBC_0.5_treatments_combined +
  theme(
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_line(),
    legend.position = c(0.002, 0.88),
    legend.justification = c("left", "top")
  ) +
  annotate("text", x = -Inf, y = Inf, label = "A. 0–5 cm depth", 
           hjust = -0.1, vjust = 1.5, size = 6, fontface = "plain") +
  scale_x_discrete(labels = site_labels)

# Remove both axis titles and legend from bottom plot; add label
p_bottom <- p_MBC_5.10_treatments_combined +
  theme(
    axis.title.y = element_blank(),
    axis.title.x = element_text(size = 16),
    legend.position = "none"
  ) +
  annotate("text", x = -Inf, y = Inf, label = "B. 5–10 cm depth", 
           hjust = -0.1, vjust = 1.5, size = 6, fontface = "plain") +
  scale_x_discrete(labels = site_labels)

# Combine vertically
combined_plot <- plot_grid(
  p_top, p_bottom,
  ncol = 1,
  align = "v",
  axis = "lr",
  rel_heights = c(1, 1)
)

# Draw final plot with shared y-axis label
# Add left margin space using a blank spacer
final_plot <- ggdraw() +
  draw_label("Microbial Biomass (mg DOC kg-1 soil)", 
             x = 0.02, y = 0.5, angle = 90, size = 18, fontface = "plain") +
  draw_plot(combined_plot, x = 0.05, y = 0, width = 0.95, height = 1)

# Display the final plot
final_plot

ggsave(
  filename = here("figures", "MBC_treatments_combined_depth_panels.png"),  
  plot = final_plot,
  width = 12,       # width in inches
  height = 6,       # height in inches
  dpi = 300         # high-resolution PNG
)
```
