---
title: "SoilMoistureAnalysis"
author: "Heather Throop"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

## Overview

This is a compilation and analysis of soil moisture data for Brittney Monus's thesis work in the GRaMPS project. Soil moisture data were provided to Brittney from GRaMPS.

This code was revised by Heather Throop in July 2025, based on the code "SoilMoisture_HT.R". The code is streamlined to avoid manual subsetting into sites and recombining sites into one file after site-specific calculations applied. In addition, this code replaces negative or 0 values for VWC with a very small number (0.0001). These are assumed to be erroneous values where the sensor is not functioning or not calibrated properly. Replacement of these erroneous values with very low values affects mean values in that 1) mean VWC is not erroneously depressed by negative numbers and 2) SWP values are not discarded during very low moisture conditions (in calculating SWP, 0 values convert to -Inf and negative values concert to NaN).

Daily moisture means are changed to be the mean of plot means (there are \~24 daily measurements be plot).

Last Update\
2027-07-10 Heather Throop

## Preparation Steps

```{r}
#| label: load packages
#| echo: false

library(tidyverse)
library(here)
```

```{r}
#| label: read in and clean data

soilmetdata = read.csv(here("data","SoilMoisture.csv")) # original file name in archive is SoilMoistureData_Request_Monus_HT.csv"

# The variable "Measure" codes for several different measured variables (VWC, T, Precip, RH). VWC is volumetric water content (VWC) and units are m3/m3

df_soilmoisture <- soilmetdata |>
  filter(Measure == "VWC") |> # Remove non VWC data
  filter(Depth.Code == "Shallow") |> # Retain only 0-10 cm depth data 
  select("site","date","Measurement","Plot","Treatment") |>
  mutate(Measurement = if_else(Measurement <= 0, 0.0001, Measurement)) # change 0 or negative VWC values to a very small positive number

```

## Convert VWC to SWP

```{r}
#| label: VWC-SWP Conversions

#need to convert VWC to soil water potential (SWP) to account for different soil 
#textures across sites - see Seth's email with code from John Bradford
#function John made comes from Crosby et al 1984
#potentially do this before calculating mean VWC and just calculate mean SWP
#exact texture for each site in the google drive 

# Function to convert VWC to SWP
VWCtoSWP_simple <- function(vwc, sand, clay){
  bar_toMPa = -0.1
  bar_conversion = 1024
  thetas <- -14.2 * sand - 3.7 * clay + 50.5
  psis <- 10 ^ (-1.58 * sand - 0.63 * clay + 2.17)
  b <- -0.3 * sand + 15.7 * clay + 3.10
  res <- psis / ((vwc * 100 / thetas) ^ b * bar_conversion) * bar_toMPa
  return(res)
}

# Create look-up table of mean sand and clay values for each site
# soil information > soil texture soil pits 2015 > John Bradford Texture 2015 submission 
soil_props <- tribble(
  ~site,        ~sand,     ~clay,
  "Antelope",   0.6464444, 0.1262222,
  "Arboretum",  0.3287,    0.2883,
  "BlackPoint", 0.7760833, 0.1108,
  "BlueChute",  0.3177,    0.3261,
  "CampColton", 0.4637778, 0.2058889
)

# Join soil properties to the measurement data
df_soilmoisture <- df_soilmoisture |>
  left_join(soil_props, by = "site") |>
  mutate(SWP = VWCtoSWP_simple(Measurement, sand, clay))

# Calculate SWP and VWC means per site-date-plot-treatment
dailymoisturemeans_byplot <- df_soilmoisture |>
  group_by(site, date, Plot, Treatment) |>
  summarise(
    N    = sum(!is.na(Measurement)),
    mean_SWP = mean(SWP, na.rm = TRUE),
    sd_SWP = sd(SWP, na.rm = TRUE),
    se_SWP = sd_SWP / sqrt(N),
    mean_VWC = mean(Measurement, na.rm = TRUE),
    sd_VWC = sd(Measurement, na.rm = TRUE),
    se_VWC = sd_VWC / sqrt(N),
    .groups = "drop"
  ) 

# Calculate SWP and VWC summary per site-date-treatment (mean of the plot means)
dailymoisturemeans <- dailymoisturemeans_byplot |>
  group_by(site, date, Treatment) |>
  summarise(
    N    = sum(!is.na(mean_SWP)),
    SWP = mean(mean_SWP, na.rm = TRUE),
    sd_SWP = sd(mean_SWP, na.rm = TRUE),
    se_SWP = sd_SWP / sqrt(N),
    VWC = mean(mean_VWC, na.rm = TRUE),
    sd_VWC = sd(mean_VWC, na.rm = TRUE),
    se_VWC = sd_VWC / sqrt(N),
    .groups = "drop"
  ) 

```

```{r}
#| label: save output

write_csv(dailymoisturemeans, here("results","dailymoisturemeans.csv"))

```

## Plot VWC Means

```{r}

ggplot(dailymoisturemeans, aes(x = date, y = VWC, color = Treatment)) +
  geom_line() +
  facet_grid( ~ site, scales = "free_y") +
  labs(x = "Date", y = "Value", title = "Mean Volumetric Water Content") +
  theme_minimal()

```
